---
title: "Proteomics Filtering Pipeline"
subtitle: "Lung Tumour — Satpathy 2021"
author: "Yashna Paul"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: hide
    theme: flatly
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 8,
  fig.height = 6,
  fig.align  = "center",
  cache      = FALSE
)
```

# Setup

> **Aim:** Load required packages and set seed for reproducibility.

```{r libraries}
suppressPackageStartupMessages({
  library(tidyverse)
  library(matrixStats)
  library(knitr)
  library(kableExtra)
})

set.seed(123)
```

# Load Data

> **Aim:** Read the raw protein and mRNA expression matrices for the Gilette
> 2020 lung tumour cohort. Rows are genes (identified by gene symbol), columns
> are samples. These are log-ratio TMT values that have already been
> centred/normalised upstream. We record the initial dimensions before any
> filtering.

```{r load-data}
data_dir <- "/cellfile/DATASETS/Human/Multi-tissue/Multi_omics/Public/LamanTripEtAl2025_processedMatrices/"

prot_data <- read_csv(paste0(data_dir, "Data/lung_satpathy2021_prot_tumor.csv"), show_col_types = FALSE)
mrna_data <- read_csv(paste0(data_dir,"Data/lung_satpathy2021_rna_tumor_idfix_subsetted.csv"), show_col_types = FALSE)

expr <- prot_data %>%
  dplyr::select(where(is.numeric)) %>%
  as.matrix()
rownames(expr) <- prot_data$symbol

mrna <- mrna_data %>%
  dplyr::select(where(is.numeric)) %>%
  as.matrix()
rownames(mrna) <- mrna_data$symbol

cat("Initial proteins:", nrow(expr), "\n")
cat("Initial mRNA genes:", nrow(mrna), "\n")
cat("Protein samples:", ncol(expr), "\n")
cat("mRNA samples:", ncol(mrna), "\n")
```

# mRNA–Protein Correlation (Before Filtering)

> **Aim:** Before applying any filters, visualise the global relationship
> between mean mRNA and mean protein expression across genes. Each dot is one
> gene, plotted at its mean expression across all samples. This serves as a
> baseline — the correlation should improve after filtering removes noisy,
> low-abundance proteins near the detection limit.

```{r cor-before-filter, fig.height=6, fig.width=12}
common_genes_raw   <- intersect(rownames(expr), rownames(mrna))
common_samples_raw <- intersect(colnames(expr), colnames(mrna))

prot_common_raw <- expr[common_genes_raw, common_samples_raw, drop = FALSE]
mrna_common_raw <- mrna[common_genes_raw, common_samples_raw, drop = FALSE]

gene_means_before <- tibble(
  gene         = common_genes_raw,
  protein_mean = rowMeans(prot_common_raw, na.rm = TRUE),
  mrna_mean    = rowMeans(mrna_common_raw, na.rm = TRUE)
) %>%
  filter(is.finite(protein_mean) & is.finite(mrna_mean))

cor_pearson_before  <- cor.test(gene_means_before$mrna_mean,
                                gene_means_before$protein_mean,
                                method = "pearson")
cor_spearman_before <- cor.test(gene_means_before$mrna_mean,
                                gene_means_before$protein_mean,
                                method = "spearman")

p_bef_pear <- ggplot(gene_means_before, aes(x = mrna_mean, y = protein_mean)) +
  geom_point(alpha = 0.3, size = 0.8, colour = "grey40") +
  geom_smooth(method = "lm", se = FALSE, colour = "red3", linewidth = 0.8) +
  labs(x = "Mean mRNA expression", y = "Mean protein expression",
       title = "Before Filtering — Pearson",
       subtitle = sprintf("r = %.3f (p = %.2e), n = %d genes",
                          cor_pearson_before$estimate,
                          cor_pearson_before$p.value,
                          nrow(gene_means_before))) +
  theme_minimal(base_size = 13)

p_bef_spear <- ggplot(gene_means_before, aes(x = mrna_mean, y = protein_mean)) +
  geom_point(alpha = 0.3, size = 0.8, colour = "grey40") +
  geom_smooth(method = "lm", se = FALSE, colour = "steelblue", linewidth = 0.8) +
  labs(x = "Mean mRNA expression", y = "Mean protein expression",
       title = "Before Filtering — Spearman",
       subtitle = sprintf("rho = %.3f (p = %.2e), n = %d genes",
                          cor_spearman_before$estimate,
                          cor_spearman_before$p.value,
                          nrow(gene_means_before))) +
  theme_minimal(base_size = 13)

library(patchwork)
p_bef_pear + p_bef_spear +
  plot_annotation(title = "Per-Gene mRNA vs Protein (Before Filtering)",
                  subtitle = "Each dot = one gene at its mean expression across all samples")
```

## Per-Gene Correlation Across Samples (Before Filtering)

> **Aim:** For each gene, compute the Spearman correlation between its protein
> and mRNA values across all samples. This gives one ρ per gene and answers:
> "For this specific gene, do samples with higher mRNA also have higher
> protein?" This is complementary to the mean-expression plot above — that
> one asks whether highly-transcribed genes tend to have high protein (across
> genes), while this one asks whether within-gene variation tracks between
> mRNA and protein (across samples). Computing this before filtering provides
> a baseline to compare against the post-filtering histogram.

```{r per-gene-cor-before, fig.height=6}
gene_cor_before <- vapply(seq_len(nrow(prot_common_raw)), function(i) {
  cor(prot_common_raw[i, ], mrna_common_raw[i, ],
      method = "spearman", use = "pairwise.complete.obs")
}, numeric(1))
names(gene_cor_before) <- rownames(prot_common_raw)
gene_cor_before <- gene_cor_before[is.finite(gene_cor_before)]
med_cor_before  <- median(gene_cor_before, na.rm = TRUE)

par(mar = c(4.2, 4.2, 1.5, 1.0), mgp = c(2.4, 0.8, 0), tcl = -0.25,
    cex.lab = 1.4, cex.axis = 1.0, bty = "l")
hist(gene_cor_before, breaks = 50, col = "grey85", border = "grey70",
     main = "", xlab = "Per-gene Spearman rho (across samples)",
     ylab = "Number of genes")
abline(v = 0, col = "black", lwd = 2, lty = 3)
abline(v = med_cor_before, col = "red3", lwd = 2.5, lty = 2)
legend("topright",
       legend = c("No correlation", sprintf("Median = %.2f", med_cor_before)),
       col = c("black", "red3"), lwd = c(2, 2.5), lty = c(3, 2),
       bty = "n", cex = 1.2, inset = 0.02)

cat(sprintf("Before filtering: %d genes, median Spearman rho = %.3f\n",
            length(gene_cor_before), med_cor_before))
```


# Filtering Pipeline

> **Aim:** Sequentially apply four filters to reduce technical noise and
> retain only features with reliable, variable measurements. The pipeline
> reduces ~11,000 proteins to ~5,000 informative features.

## Step 1: Presence Filter

> **What it does:** Computes a global expression threshold (5th percentile of
> all protein values). For each protein, calculates the fraction of samples
> where expression exceeds this threshold. Proteins detected in fewer than 95%
> of samples are removed.
>
> **Why:** Proteins near the detection limit have unreliable measurements.
> Their apparent variability may reflect missingness patterns rather than
> genuine biology. This is the most aggressive filter and typically removes
> ~40–50% of proteins.

```{r step1-presence, fig.height=10}
expr_q05 <- quantile(expr, 0.05, na.rm = TRUE)
presence <- rowMeans(expr > expr_q05, na.rm = TRUE)

par(mfrow = c(2, 1), mar = c(4.2, 4.2, 1.5, 1.0), mgp = c(2.4, 0.8, 0),
    cex.lab = 1.2, cex.axis = 1.0, bty = "l")

hist(as.vector(expr), breaks = 500, xlim = c(-10, 10),
     col = "grey85", border = "grey70",
     main = "", xlab = "Expression value", ylab = "Frequency")
abline(v = expr_q05, col = "red3", lwd = 2.5, lty = 2)
legend("topright", legend = sprintf("5%% threshold = %.2f", expr_q05),
       col = "red3", lwd = 2.5, lty = 2, bty = "n", cex = 1.1)

hist(presence, breaks = 100, xlim = c(0, 1),
     col = "grey85", border = "grey70",
     main = "", xlab = "Fraction of samples above threshold",
     ylab = "Number of proteins")
abline(v = 0.95, col = "red3", lwd = 2.5, lty = 2)
legend("topright", legend = "Presence cutoff = 0.95",
       col = "red3", lwd = 2.5, lty = 2, bty = "n", cex = 1.1)

par(mfrow = c(1, 1))

expr_filt <- expr[presence >= 0.95, , drop = FALSE]
n_after_presence <- nrow(expr_filt)
cat(sprintf("After presence filter: %d → %d proteins\n", nrow(expr), n_after_presence))
```

## Step 2: IQR Filter (Protein)

> **What it does:** Computes the interquartile range (IQR) of each protein
> across samples and removes the bottom 10%.
>
> **Why:** Proteins with near-zero IQR have essentially constant expression
> and carry no discriminative signal for downstream modelling. IQR is preferred
> over standard deviation because it is robust to outliers — a single extreme
> sample will not inflate the spread estimate and rescue a genuinely flat
> protein.

```{r step2-iqr-prot, fig.height=5}
iqr_vals <- apply(expr_filt, 1, IQR, na.rm = TRUE)
iqr_cut  <- quantile(iqr_vals, 0.10, na.rm = TRUE)

par(mar = c(4.2, 4.2, 1.5, 1.0), mgp = c(2.4, 0.8, 0),
    cex.lab = 1.2, cex.axis = 1.0, bty = "l")
hist(iqr_vals, breaks = 100,
     col = "grey85", border = "grey70",
     main = "", xlab = "IQR across samples", ylab = "Number of proteins")
abline(v = iqr_cut, col = "red3", lwd = 2.5, lty = 2)
legend("topright", legend = sprintf("IQR cutoff (10%%) = %.2f", iqr_cut),
       col = "red3", lwd = 2.5, lty = 2, bty = "n", cex = 1.1)

n_before <- nrow(expr_filt)
expr_filt <- expr_filt[iqr_vals > iqr_cut, , drop = FALSE]
n_after_iqr <- nrow(expr_filt)
cat(sprintf("After IQR filter: %d → %d proteins\n", n_before, n_after_iqr))
```

## Step 3: MAD Filter (Protein)

> **What it does:** Computes the median absolute deviation (MAD) of each
> protein across samples and removes the bottom 5%.
>
> **Why:** This is a second robustness layer. It catches proteins that passed
> the IQR filter but still have very low dispersion around the median — for
> instance, proteins with a tight IQR but a single extreme outlier inflating
> the range. This conservative cleanup removes very few additional proteins.

```{r step3-mad, fig.height=5}
mad_vals <- rowMads(expr_filt, na.rm = TRUE)
mad_cut  <- quantile(mad_vals, 0.05, na.rm = TRUE)

par(mar = c(4.2, 4.2, 1.5, 1.0), mgp = c(2.4, 0.8, 0),
    cex.lab = 1.2, cex.axis = 1.0, bty = "l")
hist(mad_vals, breaks = 100,
     col = "grey85", border = "grey70",
     main = "", xlab = "MAD across samples", ylab = "Number of proteins")
abline(v = mad_cut, col = "red3", lwd = 2.5, lty = 2)
legend("topright", legend = sprintf("MAD cutoff (5%%) = %.2f", mad_cut),
       col = "red3", lwd = 2.5, lty = 2, bty = "n", cex = 1.1)

n_before <- nrow(expr_filt)
expr_filt <- expr_filt[mad_vals > mad_cut, , drop = FALSE]
n_after_mad <- nrow(expr_filt)
cat(sprintf("After MAD filter: %d → %d proteins\n", n_before, n_after_mad))
```

## Step 4: IQR Filter (mRNA)

> **What it does:** Computes the IQR of each mRNA feature across samples and
> removes the bottom 5%.
>
> **Why:** Ensures that mRNA features used in downstream modelling (both for
> OLS residualisation and as LASSO predictors in Model B) have sufficient
> cross-sample variability. A more lenient cutoff (5% vs 10% for proteins)
> is used because mRNA data typically has fewer near-zero-variance features.

```{r step4-iqr-mrna, fig.height=5}
mrna_iqr     <- apply(mrna, 1, IQR, na.rm = TRUE)
iqr_cut_mrna <- quantile(mrna_iqr, 0.05, na.rm = TRUE)

par(mar = c(4.2, 4.2, 1.5, 1.0), mgp = c(2.4, 0.8, 0),
    cex.lab = 1.2, cex.axis = 1.0, bty = "l")
hist(mrna_iqr, breaks = 100,
     col = "grey85", border = "grey70",
     main = "", xlab = "IQR across samples", ylab = "Number of genes")
abline(v = iqr_cut_mrna, col = "red3", lwd = 2.5, lty = 2)
legend("topright", legend = sprintf("IQR cutoff (5%%) = %.2f", iqr_cut_mrna),
       col = "red3", lwd = 2.5, lty = 2, bty = "n", cex = 1.1)

mrna_filt <- mrna[mrna_iqr > iqr_cut_mrna, , drop = FALSE]
n_after_mrna_iqr <- nrow(mrna_filt)
cat(sprintf("After mRNA IQR filter: %d → %d genes\n", nrow(mrna), n_after_mrna_iqr))
```

## Step 5: Match Proteins with mRNA

> **What it does:** Intersects the gene symbols and sample IDs between the
> filtered protein and mRNA matrices so that both matrices have identical
> rows (genes) and columns (samples).
>
> **Why:** Downstream analyses (OLS residualisation, LASSO modelling) require
> paired protein–mRNA measurements for each gene in each sample. Any gene
> or sample present in only one matrix must be dropped.

```{r step5-match}
common_genes   <- intersect(rownames(expr_filt), rownames(mrna_filt))
expr_filt      <- expr_filt[common_genes, , drop = FALSE]
mrna_filt      <- mrna_filt[common_genes, , drop = FALSE]

common_samples <- intersect(colnames(expr_filt), colnames(mrna_filt))
expr_filt      <- expr_filt[, common_samples, drop = FALSE]
mrna_filt      <- mrna_filt[, common_samples, drop = FALSE]

cat(sprintf("After matching: %d genes × %d samples\n",
            nrow(expr_filt), ncol(expr_filt)))
```


# mRNA–Protein Correlation (After Filtering)

> **Aim:** Repeat the global mean-expression correlation plot from Section 3,
> now using the filtered matrices. Each dot is one gene at its mean mRNA (x)
> vs mean protein (y) across samples. Comparing with the pre-filtering plot
> shows whether the filters removed noisy genes that were distorting the
> global relationship. The correlation should increase after filtering.

```{r cor-after-filter, fig.height=6, fig.width=12}
gene_means_after <- tibble(
  gene         = rownames(expr_filt),
  protein_mean = rowMeans(expr_filt, na.rm = TRUE),
  mrna_mean    = rowMeans(mrna_filt, na.rm = TRUE)
) %>%
  filter(is.finite(protein_mean) & is.finite(mrna_mean))

cor_pearson_after  <- cor.test(gene_means_after$mrna_mean,
                               gene_means_after$protein_mean,
                               method = "pearson")
cor_spearman_after <- cor.test(gene_means_after$mrna_mean,
                               gene_means_after$protein_mean,
                               method = "spearman")

p_aft_pear <- ggplot(gene_means_after, aes(x = mrna_mean, y = protein_mean)) +
  geom_point(alpha = 0.3, size = 0.8, colour = "grey40") +
  geom_smooth(method = "lm", se = FALSE, colour = "red3", linewidth = 0.8) +
  labs(x = "Mean mRNA expression", y = "Mean protein expression",
       title = "After Filtering — Pearson",
       subtitle = sprintf("r = %.3f (p = %.2e), n = %d genes",
                          cor_pearson_after$estimate,
                          cor_pearson_after$p.value,
                          nrow(gene_means_after))) +
  theme_minimal(base_size = 13)

p_aft_spear <- ggplot(gene_means_after, aes(x = mrna_mean, y = protein_mean)) +
  geom_point(alpha = 0.3, size = 0.8, colour = "grey40") +
  geom_smooth(method = "lm", se = FALSE, colour = "steelblue", linewidth = 0.8) +
  labs(x = "Mean mRNA expression", y = "Mean protein expression",
       title = "After Filtering — Spearman",
       subtitle = sprintf("rho = %.3f (p = %.2e), n = %d genes",
                          cor_spearman_after$estimate,
                          cor_spearman_after$p.value,
                          nrow(gene_means_after))) +
  theme_minimal(base_size = 13)

p_aft_pear + p_aft_spear +
  plot_annotation(title = "Per-Gene mRNA vs Protein (After Filtering)",
                  subtitle = "Each dot = one gene at its mean expression across all samples")
```


# Protein–mRNA Correlation QC (Per-Gene, Across Samples)

> **Aim:** For each gene, compute the Spearman correlation between its protein
> and mRNA values across all samples. This gives one ρ per gene and answers:
> "For this specific gene, do samples with higher mRNA also have higher
> protein?" The distribution of these per-gene correlations characterises the
> overall strength of transcriptional control in this dataset. The median ρ
> is a key QC metric — it should be moderately positive (typically 0.3–0.5
> for TMT proteomics).

```{r step6-per-gene-cor, fig.height=6}
stopifnot(identical(rownames(expr_filt), rownames(mrna_filt)))

gene_cor <- vapply(seq_len(nrow(expr_filt)), function(i) {
  cor(expr_filt[i, ], mrna_filt[i, ], method = "spearman",
      use = "pairwise.complete.obs")
}, numeric(1))
names(gene_cor) <- rownames(expr_filt)

gene_cor <- gene_cor[is.finite(gene_cor)]
med_cor  <- median(gene_cor, na.rm = TRUE)

par(mar = c(4.2, 4.2, 0.8, 1.0), mgp = c(2.4, 0.8, 0), tcl = -0.25,
    cex.lab = 1.4, cex.axis = 1.0, bty = "l")
hist(gene_cor, breaks = 50, col = "grey85", border = "grey70",
     main = "", xlab = "Protein–mRNA correlation (Spearman rho)",
     ylab = "Number of genes")
abline(v = 0, col = "black", lwd = 2, lty = 3)
abline(v = med_cor, col = "red3", lwd = 2.5, lty = 2)
legend("topright",
       legend = c("No correlation", sprintf("Median = %.2f", med_cor)),
       col = c("black", "red3"), lwd = c(2, 2.5), lty = c(3, 2),
       bty = "n", cex = 1.2, inset = 0.02)
```

## Diagnosing the Correlation Drop {.tabset}

> **Aim:** The mean-expression Spearman correlation drops from ~0.39 (before
> filtering, ~10,700 genes) to ~0.24 (after filtering, ~6,800 genes). We
> investigate whether the presence filter selectively removes genes with
> specific expression or correlation properties.

### Classify Genes: Kept vs Filtered Out

```{r classify-genes}
kept_genes     <- rownames(expr_filt)
all_genes_raw  <- rownames(prot_common_raw)
filtered_genes <- setdiff(all_genes_raw, kept_genes)

cat(sprintf("Total genes (before): %d\n", length(all_genes_raw)))
cat(sprintf("Kept (after filtering): %d\n", length(kept_genes)))
cat(sprintf("Filtered out: %d\n", length(filtered_genes)))

# ── Compute mean expression per gene for classification ──
density_df <- tibble(
  gene         = all_genes_raw,
  protein_mean = rowMeans(prot_common_raw, na.rm = TRUE),
  mrna_mean    = rowMeans(mrna_common_raw, na.rm = TRUE),
  group        = if_else(gene %in% kept_genes, "Kept", "Filtered out")
) %>%
  filter(is.finite(protein_mean) & is.finite(mrna_mean)) %>%
  mutate(group = factor(group, levels = c("Kept", "Filtered out")))
```

### Density Plot 1: Whole Dataset — mRNA vs Protein Expression

> **What this shows:** The distribution of mean mRNA and mean protein
> expression across ALL genes before any filtering. Overlaying both on the
> same plot reveals whether the two data types have similar dynamic ranges
> and whether there is a subpopulation of genes with divergent mRNA and
> protein levels.

```{r density-all, fig.height=5}
density_all <- density_df %>%
  dplyr::select(gene, protein_mean, mrna_mean) %>%
  pivot_longer(cols = c(protein_mean, mrna_mean),
               names_to = "data_type", values_to = "mean_expr") %>%
  mutate(data_type = recode(data_type,
                            protein_mean = "Protein",
                            mrna_mean    = "mRNA"))

ggplot(density_all, aes(x = mean_expr, fill = data_type, colour = data_type)) +
  geom_density(alpha = 0.3, linewidth = 0.8) +
  scale_fill_manual(values = c("Protein" = "#E64B35", "mRNA" = "#4DBBD5")) +
  scale_colour_manual(values = c("Protein" = "#E64B35", "mRNA" = "#4DBBD5")) +
  labs(x = "Mean expression (across samples)",
       y = "Density",
       fill = NULL, colour = NULL,
       title = "Whole Dataset: mRNA vs Protein Expression",
       subtitle = sprintf("All %d genes before filtering", nrow(density_df))) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

### Density Plot 2: Filtered-Out Genes — mRNA vs Protein Expression

> **What this shows:** The expression profile of the ~4,000 genes removed by
> the filtering pipeline. The key question is the relative position of the
> mRNA and protein densities. If both shift left equally (low mRNA, low
> protein), these are simply lowly-expressed genes. If protein shifts left
> but mRNA stays central or shifts right (high mRNA, low protein), these
> genes are candidates for post-transcriptional suppression — and the filter
> is removing exactly the genes we might care about most.

```{r density-filtered, fig.height=5}
density_filt_out <- density_df %>%
  filter(group == "Filtered out") %>%
  dplyr::select(gene, protein_mean, mrna_mean) %>%
  pivot_longer(cols = c(protein_mean, mrna_mean),
               names_to = "data_type", values_to = "mean_expr") %>%
  mutate(data_type = recode(data_type,
                            protein_mean = "Protein",
                            mrna_mean    = "mRNA"))

ggplot(density_filt_out, aes(x = mean_expr, fill = data_type, colour = data_type)) +
  geom_density(alpha = 0.3, linewidth = 0.8) +
  scale_fill_manual(values = c("Protein" = "#E64B35", "mRNA" = "#4DBBD5")) +
  scale_colour_manual(values = c("Protein" = "#E64B35", "mRNA" = "#4DBBD5")) +
  labs(x = "Mean expression (across samples)",
       y = "Density",
       fill = NULL, colour = NULL,
       title = "Filtered-Out Genes: mRNA vs Protein Expression",
       subtitle = sprintf("%d genes removed by filtering",
                          sum(density_df$group == "Filtered out"))) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

### Density Plot 3: Kept Genes — mRNA vs Protein Expression

> **What this shows:** The expression profile of the ~6,800 genes retained
> after filtering. Comparing this with the filtered-out plot (above) reveals
> whether the kept genes have a systematically different mRNA-protein
> relationship. If the mRNA and protein densities are more overlapping here
> than in the filtered-out set, the kept genes have tighter transcriptional
> coupling (mRNA ≈ protein). If the mRNA density is wider or shifted
> differently, post-transcriptional control plays a different role.

```{r density-kept, fig.height=5}
density_kept <- density_df %>%
  filter(group == "Kept") %>%
  dplyr::select(gene, protein_mean, mrna_mean) %>%
  pivot_longer(cols = c(protein_mean, mrna_mean),
               names_to = "data_type", values_to = "mean_expr") %>%
  mutate(data_type = recode(data_type,
                            protein_mean = "Protein",
                            mrna_mean    = "mRNA"))

ggplot(density_kept, aes(x = mean_expr, fill = data_type, colour = data_type)) +
  geom_density(alpha = 0.3, linewidth = 0.8) +
  scale_fill_manual(values = c("Protein" = "#E64B35", "mRNA" = "#4DBBD5")) +
  scale_colour_manual(values = c("Protein" = "#E64B35", "mRNA" = "#4DBBD5")) +
  labs(x = "Mean expression (across samples)",
       y = "Density",
       fill = NULL, colour = NULL,
       title = "Kept Genes: mRNA vs Protein Expression",
       subtitle = sprintf("%d genes retained after filtering",
                          sum(density_df$group == "Kept"))) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

### Per-Gene Spearman Rho: Kept vs Filtered Out

> **What this shows:** Do the filtered-out genes have different mRNA-protein
> coupling than the kept genes? If filtered-out genes have higher per-gene
> rho, the filter is removing transcriptionally-controlled genes. If lower,
> the filtered-out genes are noisy. If the same, the filtering
> does not selectively act on the strength of within-gene mRNA-protein
> coupling.

```{r density-rho, fig.height=5}
rho_all <- vapply(seq_len(nrow(prot_common_raw)), function(i) {
  cor(prot_common_raw[i, ], mrna_common_raw[i, ],
      method = "spearman", use = "pairwise.complete.obs")
}, numeric(1))
names(rho_all) <- rownames(prot_common_raw)

rho_df <- tibble(
  gene  = names(rho_all),
  rho   = rho_all,
  group = if_else(gene %in% kept_genes, "Kept", "Filtered out")
) %>%
  filter(is.finite(rho)) %>%
  mutate(group = factor(group, levels = c("Kept", "Filtered out")))

med_kept <- median(rho_df$rho[rho_df$group == "Kept"], na.rm = TRUE)
med_filt <- median(rho_df$rho[rho_df$group == "Filtered out"], na.rm = TRUE)

ggplot(rho_df, aes(x = rho, fill = group, colour = group)) +
  geom_density(alpha = 0.3, linewidth = 0.8) +
  geom_vline(xintercept = med_kept, colour = "steelblue",
             linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = med_filt, colour = "#E64B35",
             linewidth = 1, linetype = "dashed") +
  scale_fill_manual(values = c("Kept" = "steelblue", "Filtered out" = "#E64B35")) +
  scale_colour_manual(values = c("Kept" = "steelblue", "Filtered out" = "#E64B35")) +
  labs(x = "Per-gene Spearman rho (protein vs mRNA across samples)",
       y = "Density",
       fill = NULL, colour = NULL,
       title = "Per-Gene mRNA-Protein Correlation: Kept vs Filtered Out",
       subtitle = sprintf("Median rho — Kept: %.3f, Filtered out: %.3f",
                          med_kept, med_filt)) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

wt_rho <- wilcox.test(rho ~ group, data = rho_df)
cat(sprintf("Wilcoxon test (rho: Kept vs Filtered out): p = %.2e\n", wt_rho$p.value))
```

### Summary Table: Kept vs Filtered Out

```{r density-summary}
density_df %>%
  group_by(group) %>%
  summarise(
    n           = n(),
    med_protein = median(protein_mean, na.rm = TRUE),
    med_mrna    = median(mrna_mean, na.rm = TRUE),
    .groups     = "drop"
  ) %>%
  left_join(
    rho_df %>% group_by(group) %>%
      summarise(med_rho = median(rho, na.rm = TRUE), .groups = "drop"),
    by = "group"
  ) %>%
  kable(digits = 3,
        col.names = c("Group", "N genes", "Median protein", "Median mRNA",
                       "Median per-gene rho"),
        caption = "Expression and correlation properties: kept vs filtered-out genes") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Permutation Test: Is the Correlation Drop Real?

> **Aim:** Test whether the drop in mean-expression Spearman correlation
> (from ~0.39 to ~0.24 after filtering) is explained by having fewer genes,
> or whether the filter selectively removes genes that anchored the
> correlation.
>
> **How it works:** We randomly sample the same number of genes as the
> post-filter set (n ≈ 6,800) from the full pre-filter gene set (n ≈ 10,700)
> — ignoring any expression-based criterion — and compute the Spearman
> correlation of mean mRNA vs mean protein across those randomly chosen
> genes. This is repeated 1,000 times to build a null distribution. The
> permutation acts on **gene identity** (which mRNA-protein pairs are
> included), not on sample labels or expression values.
>
> If the observed post-filter correlation falls within this null distribution,
> the drop is a sample-size artefact. If it falls significantly below (left
> tail), the filter is selectively removing genes that contributed
> disproportionately to the correlation.

```{r permutation-test, fig.height=5}
set.seed(42)

n_after <- sum(density_df$group == "Kept")
n_perms <- 1000

perm_cors_spearman <- numeric(n_perms)

for (i in seq_len(n_perms)) {
  idx <- sample(seq_len(nrow(gene_means_before)), size = n_after, replace = FALSE)
  sub <- gene_means_before[idx, ]
  perm_cors_spearman[i] <- cor(sub$mrna_mean, sub$protein_mean, method = "spearman")
}

obs_spearman <- cor_spearman_after$estimate
p_perm       <- mean(perm_cors_spearman <= obs_spearman)

par(mar = c(4.2, 4.2, 2, 1), cex.lab = 1.2, cex.axis = 1.0, bty = "l")
hist(perm_cors_spearman, breaks = 40, col = "grey85", border = "grey70",
     main = "Permutation Test: Spearman rho",
     xlab = "Correlation (random gene subset)",
     ylab = "# permutations")
abline(v = obs_spearman, col = "red3", lwd = 2.5, lty = 2)
abline(v = cor_spearman_before$estimate, col = "steelblue", lwd = 2, lty = 3)
legend("topleft",
       legend = c(sprintf("Observed (filtered) = %.3f", obs_spearman),
                  sprintf("Full dataset = %.3f", cor_spearman_before$estimate),
                  sprintf("Perm. median = %.3f", median(perm_cors_spearman))),
       col = c("red3", "steelblue", "grey50"),
       lwd = c(2.5, 2, 1), lty = c(2, 3, 1),
       bty = "n", cex = 0.9)

cat(sprintf("Permutation test (n = %d):\n", n_perms))
cat(sprintf("  Observed (filtered): %.3f\n", obs_spearman))
cat(sprintf("  Permutation median:  %.3f\n", median(perm_cors_spearman)))
cat(sprintf("  p(random <= observed): %.3f\n", p_perm))
```

```{r permutation-interpretation, results='asis'}
if (p_perm < 0.05) {
  cat("> **Interpretation:** The observed post-filter correlation is significantly\n")
  cat("> lower than expected from random subsets of the same size (p < 0.05).\n")
  cat("> The presence filter selectively removes genes that contributed to the\n")
  cat("> mRNA-protein correlation. The density plots above clarify whether\n")
  cat("> those genes had a specific expression profile (e.g., jointly low\n")
  cat("> mRNA and protein anchoring the bottom-left of the scatter).\n")
} else {
  cat("> **Interpretation:** The observed post-filter correlation falls within the\n")
  cat("> range of random subsets (p >= 0.05). The correlation drop is largely\n")
  cat("> explained by having fewer genes, not by biological differences between\n")
  cat("> kept and removed genes.\n")
}
```

# Filtering Summary

> **Aim:** Compile a summary table showing how many proteins/genes survive
> each stage, plus the before/after correlation comparison.

```{r filter-summary}
tribble(
  ~Stage,                      ~Proteins,          ~Genes_mRNA,
  "Raw",                       nrow(expr),         nrow(mrna),
  "After presence (≥95%)",     n_after_presence,   NA_integer_,
  "After IQR (top 90%)",       n_after_iqr,        NA_integer_,
  "After MAD (top 95%)",       n_after_mad,        NA_integer_,
  "After mRNA IQR (top 95%)",  n_after_mad,        n_after_mrna_iqr,
  "Final (matched)",           nrow(expr_filt),    nrow(expr_filt)
) %>%
  kable(caption = "Feature counts at each filtering stage") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

tribble(
  ~Stage,                  ~N_genes,                  ~Pearson_r,                        ~Spearman_rho,
  "Before filtering",      nrow(gene_means_before),   cor_pearson_before$estimate,        cor_spearman_before$estimate,
  "After filtering",       nrow(gene_means_after),    cor_pearson_after$estimate,         cor_spearman_after$estimate
) %>%
  kable(digits = 3,
        col.names = c("Stage", "N genes", "Pearson r", "Spearman rho"),
        caption = "mRNA–protein correlation (mean expression across genes): before vs after filtering") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

tribble(
  ~Stage,              ~N_genes,                   ~Median_rho,
  "Before filtering",  length(gene_cor_before),    med_cor_before,
  "After filtering",   length(gene_cor),           med_cor
) %>%
  kable(digits = 3,
        col.names = c("Stage", "N genes", "Median Spearman rho"),
        caption = "Per-gene mRNA–protein correlation (across samples): before vs after filtering") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```


# Interpretation and conclusions:

>1. Permutation test: The cross-gene correlation drop (0.395 → 0.249) is not a sample-size artefact — 0/1,000 random subsets produced a correlation that low. The filter selectively removes genes that anchored the mean-expression correlation.
2. Density plots: The filtered-out genes are low in both mRNA and protein, but critically, the protein density is sharper and more compressed near zero than the mRNA density. These jointly-low genes provided the bottom-left anchor points in the scatter plot. Removing them collapses the expression range and drops the cross-gene correlation. The kept genes show a similar mRNA-protein gap but with the low-expression tail trimmed.
3. Per-gene rho: This is the most informative result. The kept genes have higher within-gene mRNA-protein coupling (median ρ = 0.589) than the filtered-out genes (0.475). The filtered-out genes are not just low-expression — they are genuinely noisier at the within-gene level. This could reflect two things: (a) technical noise from measuring near the detection limit makes per-gene correlations unreliable, or (b) some of these genes truly have weaker transcriptional control, but you can't distinguish this from measurement noise when protein is barely detectable.
Combined interpretation: The filtering removes genes that are both lowly expressed and have weaker within-gene mRNA-protein coupling. The correlation drop from 0.395 to 0.249 is driven by two additive effects — loss of low-expression leverage points (density plots) AND preferential retention of genes with stronger per-gene coupling (rho plot). The kept genes are not more post-transcriptionally regulated — if anything, they are more transcriptionally controlled (higher rho), which is actually favourable for your LASSO framework: a stronger Step 0 OLS fit means cleaner residuals, which means the LASSO is working on genuine post-transcriptional signal rather than noise contaminated residuals.



# Save Filtered Matrices

> **Aim:** Save the final filtered and matched protein and mRNA matrices
> as RDS files for use in downstream LASSO modelling.

```{r save}
saveRDS(expr_filt, paste0(data_dir,"Data/20260413_Satpathy_prot_filt_YP.RDS"))
saveRDS(mrna_filt, paste0(data_dir,"Data/20260413_Satpathy_mrna_filt_YP.RDS"))

cat(sprintf("Saved: %d genes × %d samples\n", nrow(expr_filt), ncol(expr_filt)))
```

# Session Info

```{r session-info}
sessionInfo()
```
