---
title: "Extended Analysis: Post-Transcriptional Regulation in LSCC"
subtitle: "Satpathy 2021 — Residual-based LASSO discovery of protein regulators"
author: "Yashna Paul"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: hide
    theme: flatly
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 10,
  fig.height = 7,
  fig.align  = "center",
  cache      = TRUE
)
```


# Setup {.tabset}

> **Aim:** Load all required R packages, define file paths, and set up
> reusable helper functions (plotting theme, imputation, scaling). This
> section contains no analysis — it only prepares the environment. All file
> paths are centralised here. Edit one chunk to
> adapt the report to their directory structure.

## Libraries

```{r libraries}
suppressPackageStartupMessages({
  library(glmnet)
  library(matrixStats)
  library(tidyverse)  
  library(readxl)
  library(knitr)
  library(kableExtra)
  library(topGO)
  library(org.Hs.eg.db)
  library(pheatmap)
  library(ggrepel)
  library(viridis)
  library(RColorBrewer)
  library(patchwork)
})

set.seed(42)
```

## File Paths

> Update every path below to match your local
> directory layout. The report assumes that 01_filter_and_QC_trainingData_Satpathy.rmd has already been run and the filtered RDS matrices exist.
Filter_data.R takes the raw protein and mRNA matrices, applies sequential filters (presence ≥95%, protein IQR top 90%, protein MAD top 95%, mRNA IQR top 95%), matches common genes and samples between the two, runs a per-gene protein–mRNA correlation QC, and saves the filtered matrices as RDS files for downstream LASSO modelling.
> External validation datasets (unfiltered RDS files) must also be present.
> origin of 5'TOP genes: supplementary table 1: https://www.pnas.org/doi/10.1073/pnas.1912864117
> origin of mTOR and ERK pathway genes: L1000 database, a transcriptomic database of human cell lines subjected to short-term exposure to chemical inhibitors or activators targeting specific proteins. 
 Positive Marker Genes: Expression levels decrease in response to mTOR/ERK inhibition.
 Negative Marker Genes: Expression levels increase in response to mTOR/ERK inhibition.


```{r paths}

#--- Project directory --


# ── Raw data ──
data_dir <- "/cellfile/DATASETS/Human/Multi-tissue/Multi_omics/Public/LamanTripEtAl2025_processedMatrices/"

#data_dir <- "/cellfile/datapublic/keisele1/Projectmodule/"
raw_prot_tumor   <- file.path(data_dir,"Data/lung_satpathy2021_prot_tumor.csv")
raw_rna_tumor    <- file.path(data_dir,"Data/lung_satpathy2021_rna_tumor_idfix_subsetted.csv")
raw_prot_healthy <- file.path(data_dir,"Data/lung_satpathy2021_prot_healthy.csv")
raw_rna_healthy  <- file.path(data_dir,"Data/lung_satpathy2021_rna_healthy.csv")
metadata_path    <- file.path(data_dir,"Meta-data/metadata_satpathy.xlsx")

# ── Filtered matrices (output of Filter_data.R) ──
filt_prot_path <- file.path(data_dir,"Data/20260413_Satpathy_prot_filt_YP.RDS")
filt_mrna_path <- file.path(data_dir,"Data/20260413_Satpathy_mrna_filt_YP.RDS")

# ── External validation datasets (unfiltered, genes × samples) ──
# Each pair is a protein and mRNA matrix as RDS.
# Update names/paths to match your actual external cohorts.
ext_test_sets <- tibble(
  dataset      = c("Dong (Liver Tumour)", "Jiang (Liver Tumour)",
                    "Jiang (Liver Healthy)", "Xu (Lung Tumour)",
                    "Xu (Lung Healthy)", "Gilette (Lung Tumour)"),
  protein_path = file.path(data_dir, c("Data/liver_dong2022_prot_tumor.csv"
                                    ,  "Data/liver_jiang2019_prot_tumor.csv"
                                    ,  "Data/liver_jiang2019_prot_healthy.csv"
                                    ,  "Data/lung_xu2020_prot_tumor.csv"
                                    ,  "Data/lung_xu2020_prot_tumor.csv"
                                    ,  "Data/lung_gilette2020_prot_tumor.csv")),
  mrna_path    = file.path(data_dir, c("Data/liver_dong2022_rna_tumor_idfix_subsetted.csv"
                                   ,   "Data/liver_jiang2019_rna_tumor_idfix_subsetted.csv"
                                   ,   "Data/liver_gao2019_rna_healthy.csv"
                                   ,   "Data/lung_xu2020_rna_tumor_idfix_subsetted.csv"
                                   ,   "Data/lung_xu2020_rna_healthy.csv"
                                   ,   "Data/lung_gilette2020_rna_tumor_idfix_subsetted.csv"))
)

# ── Gene lists ──
gene_list_dir <- "/cellfile/datapublic/ypaul1/KatrinprojectModule"

top_genes_path       <- file.path(gene_list_dir, "5prime_TOP")
erk_neg_path         <- file.path(gene_list_dir, "250728_erk_negative_MW.csv")
erk_pos_path         <- file.path(gene_list_dir, "250728_erk_positive_MW.csv")
mtor_neg_path        <- file.path(gene_list_dir, "250728_mtor_negative_MW.csv")
mtor_pos_path        <- file.path(gene_list_dir, "250728_mtor_positive_MW.csv")


# ── Output directory ──
results_dir <- "/cellfile/datapublic/ypaul1/KatrinprojectModule/results_extended"
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
```

## Custom Theme & Helpers

> Defines a consistent ggplot theme used throughout, colour palettes for the
> three models and for positive/negative regulators, and utility functions
> for imputation, scaling, and histogram plotting. These are called
> repeatedly in later sections.

```{r theme-helpers}
theme_report <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor   = element_blank(),
      panel.grid.major   = element_line(linewidth = 0.3, colour = "grey90"),
      axis.title         = element_text(face = "plain"),
      legend.title       = element_text(face = "bold"),
      strip.text         = element_text(face = "bold"),
      plot.title         = element_text(face = "bold", size = base_size + 1),
      plot.subtitle      = element_text(colour = "grey40")
    )
}

model_colours <- c(
  "A: residuals ~ proteins" = "#E64B35",
  "B: residuals ~ mRNAs"    = "#4DBBD5",
  "C: permuted baseline"    = "#91D1C2"
)

sign_colours <- c("positive" = "firebrick", "negative" = "deepskyblue")

impute_with_train_means <- function(M, train_ref) {
  cm <- colMeans(train_ref, na.rm = TRUE)
  cm[!is.finite(cm)] <- 0
  for (j in seq_len(ncol(M))) {
    nas <- is.na(M[, j])
    if (any(nas)) M[nas, j] <- cm[j]
  }
  M
}

scale_with_train <- function(M, train_ref) {
  mu  <- colMeans(train_ref, na.rm = TRUE)
  sdv <- apply(train_ref, 2, sd, na.rm = TRUE)
  sdv[!is.finite(sdv) | sdv == 0] <- 1
  sweep(sweep(M, 2, mu, "-"), 2, sdv, "/")
}

hist_qc <- function(x, breaks = 100, xlab, ylab, vline = NULL, vlab = NULL,
                    fill_col = "grey85", border_col = "grey70") {
  df <- tibble(val = x)
  p <- ggplot(df, aes(x = val)) +
    geom_histogram(bins = breaks, fill = fill_col, colour = border_col,
                   linewidth = 0.2) +
    labs(x = xlab, y = ylab) +
    theme_report()
  if (!is.null(vline)) {
    p <- p +
      geom_vline(xintercept = vline, colour = "red3", linewidth = 1,
                 linetype = "dashed") +
      annotate("text", x = vline, y = Inf, label = vlab,
               vjust = 2, hjust = -0.1, colour = "red3", size = 4)
  }
  p
}
```


# Load Data

> **Aim:** Load the filtered protein and mRNA matrices produced by Katrin's
> `Filter_data.R` pipeline. These matrices have already undergone presence
> filtering (≥ 95% detection), protein IQR filtering (bottom 10% removed),
> MAD filtering (bottom 5% removed), mRNA IQR filtering (bottom 5% removed),
> and gene/sample matching. They are the starting point for all modelling.
> We also load sample-level metadata (age, sex, stage, smoking, molecular
> subtype) for PCA colouring and covariate checks.

```{r load-data}
protein_mat <- readRDS(filt_prot_path)
mrna_mat    <- readRDS(filt_mrna_path)

mrna_mat <- mrna_mat[rownames(protein_mat), colnames(protein_mat), drop = FALSE]
stopifnot(identical(dim(protein_mat), dim(mrna_mat)))

genes     <- rownames(protein_mat)
n_genes   <- length(genes)
n_samples <- ncol(protein_mat)

cat("Filtered data:", n_genes, "genes ×", n_samples, "samples\n")
```

```{r load-metadata}
meta_raw <- read_excel(metadata_path, sheet = "Table S1A")

meta <- meta_raw %>%
  transmute(
    sample_id = Sample.ID,
    type      = Type,
    age       = Age,
    sex       = Gender,
    stage     = Stage_mod,
    smoking   = Smoking.History_modified,
    ethnicity = Ethnicity_mod,
    nmf       = NMF.subtype,
    country   = Country.of.Origin
  )
```


# Data Overview

> **Aim:** Provide a concise summary of dataset dimensions at each filtering
> stage. This table documents how many proteins/genes and samples survive each
> filter, serving as a sanity check that no unexpected loss occurred. We also
> report the fraction of missing values (NAs) in both matrices, which informs
> the imputation strategy used later.

```{r data-overview}
raw_prot <- read_csv(raw_prot_tumor, show_col_types = FALSE)
raw_rna  <- read_csv(raw_rna_tumor,  show_col_types = FALSE)

tribble(
  ~Matrix,                ~Genes,            ~Samples,
  "Raw protein (tumour)", nrow(raw_prot),    ncol(raw_prot) - 1,
  "Raw mRNA (tumour)",    nrow(raw_rna),     ncol(raw_rna) - 1,
  "Filtered protein",     nrow(protein_mat), ncol(protein_mat),
  "Filtered mRNA",        nrow(mrna_mat),    ncol(mrna_mat),
  "Common (final)",       n_genes,           n_samples
) %>%
  kable(caption = "Dataset dimensions at each processing stage",
        format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

cat(sprintf("Protein NA fraction: %.3f%%\nmRNA NA fraction: %.3f%%\n",
            mean(is.na(protein_mat)) * 100, mean(is.na(mrna_mat)) * 100))

rm(raw_prot, raw_rna); gc(verbose = FALSE)
```


# Filtering QC

> **Aim:** Reproduce and visualise Katrin's four-step filtering pipeline to
> document the effect of each filter on feature count. This is not re-running
> the filter (the filtered matrices are already loaded); rather, it re-applies
> the same logic to the raw data for documentation purposes, showing the
> distribution of each metric and where the cutoff falls. This ensures
> transparency and reproducibility.

## Presence Filter

> **What it does:** Removes proteins that fall below the global 5th-percentile
> expression threshold in more than 5% of samples. Proteins that are "present"
> in fewer than 95% of samples are unreliable — their apparent variability
> may be driven by missingness patterns rather than genuine biology. This is
> the most aggressive filter and typically removes ~40–50% of the initial
> protein catalogue.

```{r presence-filter, fig.height=5}
raw_prot <- read_csv(raw_prot_tumor, show_col_types = FALSE)
expr_raw <- raw_prot %>% dplyr::select(where(is.numeric)) %>% as.matrix()
rownames(expr_raw) <- raw_prot$symbol

expr_q05 <- quantile(expr_raw, 0.05, na.rm = TRUE)
presence <- rowMeans(expr_raw > expr_q05, na.rm = TRUE)

p1 <- hist_qc(as.vector(expr_raw), breaks = 500,
               xlab = "Expression value", ylab = "Frequency",
               vline = expr_q05,
               vlab = sprintf("5%% threshold = %.2f", expr_q05)) +
  coord_cartesian(xlim = c(-10, 10))

p2 <- hist_qc(presence, breaks = 100,
               xlab = "Fraction samples above threshold", ylab = "# proteins",
               vline = 0.95, vlab = "Presence cutoff = 0.95")

p1 / p2 + plot_annotation(
  title = "Presence Filter",
  subtitle = sprintf("%d → %d proteins", nrow(expr_raw), sum(presence >= 0.95)))

expr_pres <- expr_raw[presence >= 0.95, , drop = FALSE]
```

## IQR Filter (Protein)

> **What it does:** Removes the bottom 10% of proteins by interquartile range
> (IQR). Proteins with near-zero IQR have essentially constant expression
> across samples and carry no discriminative signal for the LASSO. IQR is
> preferred over variance or standard deviation because it is robust to
> outliers — a single extreme sample will not inflate the spread estimate and
> rescue a genuinely flat protein from removal.

```{r iqr-filter-prot, fig.height=4}
iqr_vals <- apply(expr_pres, 1, IQR, na.rm = TRUE)
iqr_cut  <- quantile(iqr_vals, 0.10, na.rm = TRUE)

hist_qc(iqr_vals, xlab = "IQR across samples", ylab = "# proteins",
        vline = iqr_cut,
        vlab = sprintf("IQR cutoff (10%%) = %.2f", iqr_cut)) +
  labs(title = "Protein IQR Filter",
       subtitle = sprintf("%d → %d", nrow(expr_pres), sum(iqr_vals > iqr_cut)))

expr_iqr <- expr_pres[iqr_vals > iqr_cut, , drop = FALSE]
```

## MAD Filter (Protein)

> **What it does:** Removes the bottom 5% of proteins by median absolute
> deviation (MAD). This is a second robustness layer that catches proteins
> which passed the IQR filter but still have very low dispersion around
> the median. For example, a protein with a tight IQR but one extreme outlier
> inflating its range would pass IQR but fail MAD. This conservative cleanup
> removes very few additional proteins.

```{r mad-filter, fig.height=4}
mad_vals <- rowMads(expr_iqr, na.rm = TRUE)
mad_cut  <- quantile(mad_vals, 0.05, na.rm = TRUE)

hist_qc(mad_vals, xlab = "MAD across samples", ylab = "# proteins",
        vline = mad_cut,
        vlab = sprintf("MAD cutoff (5%%) = %.2f", mad_cut)) +
  labs(title = "Protein MAD Filter",
       subtitle = sprintf("%d → %d", nrow(expr_iqr), sum(mad_vals > mad_cut)))

rm(expr_raw, expr_pres, expr_iqr); gc(verbose = FALSE)
```

## IQR Filter (mRNA)

> **What it does:** Removes the bottom 5% of mRNA features by IQR. This
> ensures that both the OLS residualisation step (Step 0) and Model B
> (residuals ~ other mRNAs) have predictors with sufficient cross-sample
> variability. A more lenient cutoff (5% vs 10% for proteins) is used because
> mRNA data typically has fewer near-zero-variance features.

```{r iqr-filter-mrna, fig.height=4}
raw_rna  <- read_csv(raw_rna_tumor, show_col_types = FALSE)
mrna_raw <- raw_rna %>% dplyr::select(where(is.numeric)) %>% as.matrix()
rownames(mrna_raw) <- raw_rna$symbol

mrna_iqr     <- apply(mrna_raw, 1, IQR, na.rm = TRUE)
iqr_cut_mrna <- quantile(mrna_iqr, 0.05, na.rm = TRUE)

hist_qc(mrna_iqr, xlab = "IQR across samples", ylab = "# genes",
        vline = iqr_cut_mrna,
        vlab = sprintf("IQR cutoff (5%%) = %.2f", iqr_cut_mrna)) +
  labs(title = "mRNA IQR Filter",
       subtitle = sprintf("%d → %d", nrow(mrna_raw), sum(mrna_iqr > iqr_cut_mrna)))

rm(raw_prot, raw_rna, mrna_raw); gc(verbose = FALSE)
```


# Sample-Level QC {.tabset}

> **Aim:** Identify technical outlier samples before modelling. Outlier samples
> bias model training (if they fall into training folds) and inflate test-set
> error (if they fall into test folds). Two complementary approaches are used:
>
> 1. **PCA** — projects all samples into a low-dimensional space. If technical
>    variables (TMT plex, batch) dominate PC1/PC2, the data may need batch
>    correction. If biological variables (stage, subtype) dominate, the data
>    is well-behaved.
> 2. **Per-sample mRNA–protein correlation** — computes the Spearman ρ
>    between all protein and mRNA values within each sample. A normal sample
>    should show moderate positive ρ. Abnormally low ρ suggests a sample
>    swap, RNA/protein degradation, or failed measurement.

## PCA by Metadata

> **What this shows:** The scree plot reveals how many PCs carry meaningful
> variance (guiding dimensionality interpretation). The coloured scatter plots
> reveal whether biological or technical variables drive the dominant axes of
> variation. Ideally, biological variables (molecular subtype, stage) should
> separate samples, while technical variables (sex, country) should not
> produce strong clustering unless they are confounded with biology.

```{r pca-qc, fig.height=6}
mat_imp <- t(apply(protein_mat, 1, function(x) {
  x[!is.finite(x)] <- NA_real_
  x[is.na(x)] <- median(x, na.rm = TRUE)
  x
}))

pca <- prcomp(t(mat_imp), center = TRUE, scale. = FALSE)
var_expl <- (pca$sdev^2) / sum(pca$sdev^2)

pca_df <- tibble(sample_id = colnames(protein_mat),
                 PC1 = pca$x[, 1], PC2 = pca$x[, 2], PC3 = pca$x[, 3]) %>%
  left_join(meta, by = "sample_id")

ggplot(tibble(PC = 1:20, var = var_expl[1:20] * 100), aes(PC, var)) +
  geom_col(fill = "steelblue", width = 0.7) +
  geom_line(colour = "grey30") + geom_point(size = 2) +
  labs(x = "Principal Component", y = "% Variance Explained",
       title = "Scree Plot",
       subtitle = "How many PCs carry meaningful variance?") +
  theme_report()
```

```{r pca-panels, fig.height=12, fig.width=10}
vars_to_plot <- intersect(c("sex", "stage", "smoking", "nmf", "country"),
                          colnames(pca_df))

pca_plots <- lapply(vars_to_plot, function(v) {
  ggplot(pca_df %>% filter(!is.na(.data[[v]])),
         aes(PC1, PC2, colour = .data[[v]])) +
    geom_point(size = 2.5, alpha = 0.8) +
    labs(x = sprintf("PC1 (%.1f%%)", var_expl[1] * 100),
         y = sprintf("PC2 (%.1f%%)", var_expl[2] * 100),
         colour = v, title = paste("Coloured by", v)) +
    theme_report(11)
})

wrap_plots(pca_plots, ncol = 2) +
  plot_annotation(title = "PCA: Sample-Level Structure",
                  subtitle = "Filtered protein matrix — Satpathy 2021 tumour samples")
```

## Per-Sample mRNA–Protein Correlation

> **What this answers:** Is there a sample where the protein-mRNA relationship
> has broken down? For each sample, we compute the Spearman ρ between its
> protein and mRNA values across all genes. A well-behaved sample typically
> shows ρ ≈ 0.3–0.6. Samples below the Tukey outlier threshold (Q1 − 1.5×IQR)
> are flagged. Flagged samples are not automatically removed but should be
> investigated — if they cluster by a metadata variable, the problem may be
> systematic (e.g., degradation in samples from a specific site).

```{r per-sample-cor, fig.height=5}
sample_cors <- vapply(seq_len(n_samples), function(j) {
  p <- protein_mat[, j]; m <- mrna_mat[, j]
  ok <- is.finite(p) & is.finite(m)
  if (sum(ok) < 30) return(NA_real_)
  cor(p[ok], m[ok], method = "spearman")
}, numeric(1))
names(sample_cors) <- colnames(protein_mat)

sample_cor_df <- tibble(sample_id = names(sample_cors), rho = sample_cors) %>%
  left_join(meta, by = "sample_id")

med_cor <- median(sample_cors, na.rm = TRUE)
outlier_threshold <- quantile(sample_cors, 0.25, na.rm = TRUE) -
                     1.5 * IQR(sample_cors, na.rm = TRUE)

ggplot(sample_cor_df, aes(x = rho)) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "black",
                 alpha = 0.7, linewidth = 0.2) +
  geom_vline(xintercept = med_cor, colour = "red3",
             linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = outlier_threshold, colour = "orange",
             linewidth = 1, linetype = "dotted") +
  annotate("text", x = med_cor, y = Inf,
           label = sprintf("Median = %.3f", med_cor),
           vjust = 2, hjust = -0.1, colour = "red3", size = 3.5) +
  annotate("text", x = outlier_threshold, y = Inf,
           label = sprintf("Outlier cutoff = %.3f", outlier_threshold),
           vjust = 3.5, hjust = -0.1, colour = "orange", size = 3.5) +
  labs(x = "Per-sample Spearman ρ (protein vs mRNA across genes)",
       y = "# samples",
       title = "Per-Sample mRNA–Protein Correlation",
       subtitle = "Low values may indicate technical artefacts or sample swaps") +
  theme_report()
```

```{r outlier-samples}
outlier_samples <- sample_cor_df %>%
  filter(rho < outlier_threshold | is.na(rho)) %>% arrange(rho)

if (nrow(outlier_samples) > 0) {
  outlier_samples %>%
    dplyr::select(sample_id, rho, sex, stage, smoking) %>%
    kable(digits = 3, caption = "Flagged outlier samples (ρ below Tukey threshold)") %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
} else {
  cat("✓ No outlier samples detected.\n")
}
```

> **Follow-up:** Do outlier samples cluster by any metadata variable? If so,
> the issue may be technical (e.g., a specific TMT plex) rather than random.

```{r sample-cor-covariates, fig.height=5, fig.width=12}
p_cor_boxes <- lapply(c("sex", "stage", "smoking"), function(v) {
  ggplot(sample_cor_df %>% filter(!is.na(.data[[v]])),
         aes(.data[[v]], rho, fill = .data[[v]])) +
    geom_boxplot(width = 0.6, alpha = 0.7, outlier.alpha = 0.5) +
    geom_jitter(width = 0.15, size = 1, alpha = 0.4) +
    labs(x = v, y = "Spearman ρ", title = paste("By", v)) +
    theme_report(11) +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 30, hjust = 1))
})
wrap_plots(p_cor_boxes, ncol = 3)
```

## Per-Gene mRNA–Protein Correlation

> **What this answers:** How strongly does mRNA predict protein across samples,
> on a gene-by-gene basis? This distribution characterises the overall strength
> of transcriptional control in this tissue/tumour type. The median ρ tells us
> what fraction of protein variance is typically explained by mRNA. Genes in
> the left tail (low or negative ρ) are the most interesting targets for
> post-transcriptional regulation discovery — their protein levels diverge
> from what mRNA would predict.

```{r per-gene-cor, fig.height=5}
gene_cors <- vapply(seq_len(n_genes), function(i) {
  cor(protein_mat[i, ], mrna_mat[i, ], method = "spearman",
      use = "pairwise.complete.obs")
}, numeric(1))
names(gene_cors) <- genes
gene_cors <- gene_cors[is.finite(gene_cors)]
med_gene_cor <- median(gene_cors, na.rm = TRUE)

ggplot(tibble(rho = gene_cors), aes(x = rho)) +
  geom_histogram(bins = 50, fill = "grey85", colour = "grey70", linewidth = 0.2) +
  geom_vline(xintercept = 0, colour = "black", linewidth = 1, linetype = "dotted") +
  geom_vline(xintercept = med_gene_cor, colour = "red3",
             linewidth = 1, linetype = "dashed") +
  annotate("text", x = med_gene_cor, y = Inf,
           label = sprintf("Median = %.2f", med_gene_cor),
           vjust = 2, hjust = -0.1, colour = "red3", size = 4) +
  labs(x = "Per-gene Spearman ρ (protein vs mRNA across samples)",
       y = "# genes",
       title = "Per-Gene mRNA–Protein Correlation",
       subtitle = "Low-ρ genes are prime targets for post-transcriptional discovery") +
  theme_report()
```


# Step 0: Protein Residualisation

> **Aim:** Remove the transcriptional component of protein variation before
> searching for post-transcriptional regulators. For each gene, we regress
> protein expression on its own mRNA using ordinary least squares (OLS,
> unpenalised). The residuals represent the portion of protein abundance that
> **cannot** be explained by that gene's own transcription.
>
> **Why this matters:** In the original pipeline, mRNA_self was included as a
> penalised LASSO feature competing with all protein predictors. The L1 penalty
> could shrink mRNA_self to zero, meaning some targets were never corrected for
> their own mRNA at all. By separating the mRNA correction into an explicit,
> unpenalised OLS step, we guarantee that transcription is always accounted
> for. Any R² from the downstream LASSO is therefore **purely**
> post-transcriptional by construction.
>
> We also save the OLS intercepts and slopes per gene — these will be needed
> to compute residuals in external validation datasets using training-set
> statistics (preventing data leakage).

```{r residualise, fig.height=5}
residual_mat <- matrix(NA_real_, nrow = n_genes, ncol = n_samples,
                       dimnames = list(genes, colnames(protein_mat)))
ols_r2       <- setNames(rep(NA_real_, n_genes), genes)
ols_slope    <- setNames(rep(NA_real_, n_genes), genes)
ols_intercept <- setNames(rep(NA_real_, n_genes), genes)

for (i in seq_len(n_genes)) {
  p <- as.numeric(protein_mat[i, ])
  m <- as.numeric(mrna_mat[i, ])
  ok <- is.finite(p) & is.finite(m)
  if (sum(ok) < 10) next

  fit <- lm(p[ok] ~ m[ok])
  residual_mat[i, which(ok)] <- residuals(fit)
  ols_r2[i]        <- summary(fit)$r.squared
  ols_intercept[i] <- coef(fit)[1]
  ols_slope[i]     <- coef(fit)[2]
}

ols_df <- tibble(gene = genes, r2 = ols_r2, slope = ols_slope) %>%
  filter(is.finite(r2))
med_ols_r2 <- median(ols_df$r2, na.rm = TRUE)

ggplot(ols_df, aes(x = r2)) +
  geom_histogram(bins = 50, fill = "grey85", colour = "grey70", linewidth = 0.2) +
  geom_vline(xintercept = med_ols_r2, colour = "red3",
             linewidth = 1, linetype = "dashed") +
  annotate("text", x = med_ols_r2, y = Inf,
           label = sprintf("Median OLS R² = %.3f", med_ols_r2),
           vjust = 2, hjust = -0.1, colour = "red3", size = 4) +
  labs(x = "OLS R² (protein ~ own mRNA)", y = "# genes",
       title = "Step 0: Variance Explained by Own mRNA",
       subtitle = "Residuals from this OLS are the response variable for Models A, B, C") +
  theme_report()

cat(sprintf("Median OLS R²: %.3f\n", med_ols_r2))
cat(sprintf("Genes with R² > 0.3: %d (%.1f%%)\n",
            sum(ols_df$r2 > 0.3), mean(ols_df$r2 > 0.3) * 100))
cat(sprintf("Genes with R² < 0.05: %d (%.1f%%) — mostly post-transcriptionally governed\n",
            sum(ols_df$r2 < 0.05), mean(ols_df$r2 < 0.05) * 100))
```


# LASSO Modelling: Nested Cross-Validation {.tabset}

> **Aim:** Estimate out-of-sample R² and MSE for three models, all predicting
> **protein residuals** (from Step 0):
>
> - **Model A** (residuals ~ other proteins): LASSO regression using all other
>   proteins as predictors. Identifies proteins whose abundance predicts
>   post-transcriptional variation in the target.
> - **Model B** (residuals ~ other mRNAs): LASSO regression using all other
>   mRNAs as predictors. Critical control — if an mRNA predicts residuals as
>   well as its corresponding protein, the "regulation" may be co-
>   transcriptional leakage rather than genuine post-transcriptional biology.
> - **Model C** (permuted baseline): Same predictors as Model A, but target
>   residuals are shuffled. Establishes the null — any model performing no
>   better than C is fitting noise.
>
> **Nested CV structure:** The outer loop (10-fold) holds out test samples
> for unbiased performance estimation. The inner loop (5-fold, within each
> outer training set) selects the optimal regularisation parameter λ.
> `cv.glmnet` returns two candidates: `lambda.min` (lowest inner CV error)
> and `lambda.1se` (most regularised model within 1 standard error of the
> minimum). We use `lambda.1se` because it favours sparser models, reducing
> false positive regulators at the cost of slightly higher error — the right
> trade-off for a discovery-oriented analysis.

## Parameters

```{r lasso-params}
outer_k <- 10
inner_k <- 5
min_n_train <- 15

# High-confidence regulator thresholds
hc_min_folds    <- 8     # selected in ≥ 8 of 10 outer folds
hc_min_targets  <- 50    # predicts ≥ 50 target residuals
hc_min_strength <- 0.5   # |frac_positive − frac_negative| ≥ 0.5
```

## Run Nested CV

> **What happens here:** For each of the 10 outer folds, we: (1) split samples
> into train/test, (2) build scaled predictor matrices for both proteins and
> mRNAs using training-set statistics only, (3) for each target gene, fit
> Models A, B, and C on the training residuals using inner 5-fold CV to select
> lambda, (4) predict test-set residuals and compute MSE. We also record all
> non-zero coefficients and the selected lambda for diagnostic plotting.

```{r lasso-run, cache=TRUE}
fold_id <- sample(rep(seq_len(outer_k), length.out = n_samples))

mse_folds_A <- mse_folds_B <- mse_folds_C <- vector("list", outer_k)
coef_A_list <- coef_B_list <- vector("list", outer_k)
sel_counts_A <- sel_counts_B <- vector("list", outer_k)
lambda_selected_A <- lambda_selected_B <- vector("list", outer_k)

example_cvfit_A <- example_cvfit_B <- NULL
example_target  <- NULL

cat("Running nested CV (", outer_k, "outer ×", inner_k, "inner folds)...\n")

for (f in seq_len(outer_k)) {
  cat("  Outer fold", f, "/", outer_k, "\n")

  test_idx  <- which(fold_id == f)
  train_idx <- which(fold_id != f)
  n_tr <- length(train_idx)

  # ── Protein predictors ──
  X_prot_tr_raw <- t(protein_mat[, train_idx, drop = FALSE])
  X_prot_te_raw <- t(protein_mat[, test_idx,  drop = FALSE])
  X_prot_tr_imp <- impute_with_train_means(X_prot_tr_raw, X_prot_tr_raw)
  X_prot_te_imp <- impute_with_train_means(X_prot_te_raw, X_prot_tr_raw)
  X_prot_tr <- scale_with_train(X_prot_tr_imp, X_prot_tr_imp)
  X_prot_te <- scale_with_train(X_prot_te_imp, X_prot_tr_imp)
  storage.mode(X_prot_tr) <- storage.mode(X_prot_te) <- "double"

  vx <- apply(X_prot_tr, 2, var, na.rm = TRUE)
  keep_p <- is.finite(vx) & vx > 0
  X_prot_tr <- X_prot_tr[, keep_p, drop = FALSE]
  X_prot_te <- X_prot_te[, keep_p, drop = FALSE]
  prot_preds <- colnames(X_prot_tr)

  # ── mRNA predictors ──
  X_mrna_tr_raw <- t(mrna_mat[, train_idx, drop = FALSE])
  X_mrna_te_raw <- t(mrna_mat[, test_idx,  drop = FALSE])
  X_mrna_tr_imp <- impute_with_train_means(X_mrna_tr_raw, X_mrna_tr_raw)
  X_mrna_te_imp <- impute_with_train_means(X_mrna_te_raw, X_mrna_tr_raw)
  X_mrna_tr <- scale_with_train(X_mrna_tr_imp, X_mrna_tr_imp)
  X_mrna_te <- scale_with_train(X_mrna_te_imp, X_mrna_tr_imp)
  storage.mode(X_mrna_tr) <- storage.mode(X_mrna_te) <- "double"

  vx_m <- apply(X_mrna_tr, 2, var, na.rm = TRUE)
  keep_m <- is.finite(vx_m) & vx_m > 0
  X_mrna_tr <- X_mrna_tr[, keep_m, drop = FALSE]
  X_mrna_te <- X_mrna_te[, keep_m, drop = FALSE]
  mrna_preds <- colnames(X_mrna_tr)

  # ── Per-fold containers ──
  sel_A <- setNames(integer(length(prot_preds)), prot_preds)
  sel_B <- setNames(integer(length(mrna_preds)), mrna_preds)
  cA <- list(); cB <- list(); kA <- 1; kB <- 1

  mse_A <- mse_B <- mse_C <- setNames(rep(NA_real_, n_genes), genes)
  lam_A_vec <- lam_B_vec <- setNames(rep(NA_real_, n_genes), genes)

  for (target in genes) {
    y_tr <- as.numeric(residual_mat[target, train_idx])
    y_te <- as.numeric(residual_mat[target, test_idx])
    ok_tr <- is.finite(y_tr); ok_te <- is.finite(y_te)
    if (sum(ok_tr) < min_n_train || sum(ok_te) < 2) next

    fid <- sample(rep(seq_len(inner_k), length.out = sum(ok_tr)))

    # ── Model A: residuals ~ proteins ──
    cur_p <- if (target %in% prot_preds) prot_preds[prot_preds != target] else prot_preds
    if (length(cur_p) < 2) next

    cv_A <- cv.glmnet(X_prot_tr[ok_tr, cur_p, drop = FALSE], y_tr[ok_tr],
                      alpha = 1, nfolds = inner_k, foldid = fid,
                      standardize = FALSE, family = "gaussian")
    lam_A <- cv_A$lambda.1se; lam_A_vec[target] <- lam_A

    if (is.null(example_cvfit_A) && target == genes[min(100, n_genes)]) {
      example_cvfit_A <- cv_A; example_cvfit_B_placeholder <- TRUE
      example_target <- target
    }

    yhat_A <- as.numeric(predict(cv_A, newx = X_prot_te[ok_te, cur_p, drop = FALSE], s = lam_A))
    mse_A[target] <- mean((y_te[ok_te] - yhat_A)^2)

    b_A <- as.matrix(coef(cv_A, s = lam_A))[-1, , drop = FALSE]
    act_A <- which(b_A[, 1] != 0)
    if (length(act_A) > 0) {
      an <- rownames(b_A)[act_A]
      sel_A[intersect(an, names(sel_A))] <- sel_A[intersect(an, names(sel_A))] + 1L
      cA[[kA]] <- tibble(fold = f, target = target, regulator = an,
                         beta = b_A[act_A, 1],
                         sign = if_else(b_A[act_A, 1] > 0, "positive", "negative"))
      kA <- kA + 1
    }

    # ── Model B: residuals ~ mRNAs ──
    cur_m <- if (target %in% mrna_preds) mrna_preds[mrna_preds != target] else mrna_preds
    if (length(cur_m) < 2) next

    cv_B <- cv.glmnet(X_mrna_tr[ok_tr, cur_m, drop = FALSE], y_tr[ok_tr],
                      alpha = 1, nfolds = inner_k, foldid = fid,
                      standardize = FALSE, family = "gaussian")
    lam_B <- cv_B$lambda.1se; lam_B_vec[target] <- lam_B

    if (!is.null(example_cvfit_A) && is.null(example_cvfit_B) && target == example_target) {
      example_cvfit_B <- cv_B
    }

    yhat_B <- as.numeric(predict(cv_B, newx = X_mrna_te[ok_te, cur_m, drop = FALSE], s = lam_B))
    mse_B[target] <- mean((y_te[ok_te] - yhat_B)^2)

    b_B <- as.matrix(coef(cv_B, s = lam_B))[-1, , drop = FALSE]
    act_B <- which(b_B[, 1] != 0)
    if (length(act_B) > 0) {
      an_B <- rownames(b_B)[act_B]
      sel_B[intersect(an_B, names(sel_B))] <- sel_B[intersect(an_B, names(sel_B))] + 1L
      cB[[kB]] <- tibble(fold = f, target = target, regulator = an_B,
                         beta = b_B[act_B, 1],
                         sign = if_else(b_B[act_B, 1] > 0, "positive", "negative"))
      kB <- kB + 1
    }

    # ── Model C: permuted residuals ~ proteins ──
    y_perm <- sample(y_tr[ok_tr], replace = FALSE)
    cv_C <- cv.glmnet(X_prot_tr[ok_tr, cur_p, drop = FALSE], y_perm,
                      alpha = 1, nfolds = inner_k, foldid = fid,
                      standardize = FALSE, family = "gaussian")
    yhat_C <- as.numeric(predict(cv_C, newx = X_prot_te[ok_te, cur_p, drop = FALSE],
                                  s = cv_C$lambda.1se))
    mse_C[target] <- mean((y_te[ok_te] - yhat_C)^2)
  }

  mse_folds_A[[f]] <- mse_A; mse_folds_B[[f]] <- mse_B; mse_folds_C[[f]] <- mse_C
  sel_counts_A[[f]] <- sel_A; sel_counts_B[[f]] <- sel_B
  coef_A_list[[f]] <- bind_rows(cA); coef_B_list[[f]] <- bind_rows(cB)
  lambda_selected_A[[f]] <- lam_A_vec; lambda_selected_B[[f]] <- lam_B_vec

  cat("    Targets — A:", sum(is.finite(mse_A)),
      " B:", sum(is.finite(mse_B)), " C:", sum(is.finite(mse_C)), "\n")
}
cat("Nested CV complete.\n")
```

## Aggregate Results

> **What this does:** Converts the per-fold MSE lists into matrices
> (fold × target), computes mean MSE and R² per target, and aggregates
> non-zero coefficient counts across folds for regulator ranking.

```{r aggregate}
make_mse_mat <- function(ml) {
  mat <- do.call(rbind, lapply(ml, as.numeric))
  colnames(mat) <- genes; rownames(mat) <- paste0("fold", seq_along(ml)); mat
}
mse_mat_A <- make_mse_mat(mse_folds_A)
mse_mat_B <- make_mse_mat(mse_folds_B)
mse_mat_C <- make_mse_mat(mse_folds_C)

mean_mse_A <- colMeans(mse_mat_A, na.rm = TRUE)
mean_mse_B <- colMeans(mse_mat_B, na.rm = TRUE)
mean_mse_C <- colMeans(mse_mat_C, na.rm = TRUE)

resid_var <- apply(residual_mat, 1, var, na.rm = TRUE)
r2_A <- 1 - mean_mse_A / resid_var[genes]
r2_B <- 1 - mean_mse_B / resid_var[genes]
r2_C <- 1 - mean_mse_C / resid_var[genes]

r2_df <- bind_rows(
  tibble(target = genes, r2 = r2_A, model = "A: residuals ~ proteins"),
  tibble(target = genes, r2 = r2_B, model = "B: residuals ~ mRNAs"),
  tibble(target = genes, r2 = r2_C, model = "C: permuted baseline")
) %>% filter(is.finite(r2)) %>%
  mutate(model = factor(model, levels = names(model_colours)))

coef_A_df <- bind_rows(coef_A_list)
coef_B_df <- bind_rows(coef_B_list)

agg_sel <- function(sl) {
  nms <- sort(unique(unlist(lapply(sl, names))))
  mat <- matrix(0L, nrow = outer_k, ncol = length(nms),
                dimnames = list(paste0("fold", seq_len(outer_k)), nms))
  for (f in seq_along(sl)) { sc <- sl[[f]]; mat[f, names(sc)] <- as.integer(sc) }
  mat
}
sel_mat_A <- agg_sel(sel_counts_A)
sel_mat_B <- agg_sel(sel_counts_B)
```


# Lambda Diagnostics {.tabset}

> **Aim:** Verify that the inner CV is selecting λ sensibly. Three diagnostic
> plots address three questions:
>
> 1. **Lambda path** — Is there a clear minimum in CV error? A flat curve
>    means the data lacks signal for this target.
> 2. **Nonzero coefficients vs λ** — Does sparsity increase monotonically
>    with λ? Non-monotonicity would indicate a `glmnet` pathology.
> 3. **Lambda distribution across targets** — Is regularisation strength
>    consistent? A bimodal distribution suggests two regimes (some targets
>    with strong signal, others near-null).

## Lambda Path (Example Target)

> Shows inner CV error as a function of log(λ) for one representative target.
> The left dashed line is `lambda.min` (lowest error); the right dashed line
> is `lambda.1se` (our choice — most regularised within 1 SE of minimum).
> A well-behaved curve has a clear U-shape.

```{r lambda-path, fig.height=10}
par(mfrow = c(2, 1))
if (!is.null(example_cvfit_A))
  plot(example_cvfit_A, main = sprintf("Model A — target: %s", example_target))
if (!is.null(example_cvfit_B))
  plot(example_cvfit_B, main = sprintf("Model B — target: %s", example_target))
par(mfrow = c(1, 1))
```

## Nonzero Coefficients vs Lambda

> Shows how the number of selected predictors decreases as λ increases.
> Should be monotonically non-increasing. The vertical lines mark the two
> candidate lambdas. The gap between `lambda.min` and `lambda.1se` tells you
> how many features the conservative choice sacrifices for parsimony.

```{r nzero-vs-lambda, fig.height=10}
if (!is.null(example_cvfit_A)) {
  gl_A <- example_cvfit_A$glmnet.fit
  p1 <- ggplot(tibble(ll = log(gl_A$lambda), nz = as.numeric(gl_A$df)),
               aes(ll, nz)) +
    geom_line(colour = "steelblue", linewidth = 0.8) +
    geom_vline(xintercept = log(example_cvfit_A$lambda.1se),
               colour = "red3", linetype = "dashed") +
    geom_vline(xintercept = log(example_cvfit_A$lambda.min),
               colour = "orange", linetype = "dashed") +
    annotate("text", x = log(example_cvfit_A$lambda.1se), y = Inf,
             label = "lambda.1se", vjust = 2, hjust = -0.1, colour = "red3", size = 3) +
    annotate("text", x = log(example_cvfit_A$lambda.min), y = Inf,
             label = "lambda.min", vjust = 3.5, hjust = -0.1, colour = "orange", size = 3) +
    labs(x = "log(λ)", y = "# nonzero coefficients",
         title = sprintf("Model A — %s", example_target)) +
    theme_report()

  gl_B <- example_cvfit_B$glmnet.fit
  p2 <- ggplot(tibble(ll = log(gl_B$lambda), nz = as.numeric(gl_B$df)),
               aes(ll, nz)) +
    geom_line(colour = "#4DBBD5", linewidth = 0.8) +
    geom_vline(xintercept = log(example_cvfit_B$lambda.1se),
               colour = "red3", linetype = "dashed") +
    geom_vline(xintercept = log(example_cvfit_B$lambda.min),
               colour = "orange", linetype = "dashed") +
    labs(x = "log(λ)", y = "# nonzero coefficients",
         title = sprintf("Model B — %s", example_target)) +
    theme_report()

  p1 / p2
}
```

## Lambda Distribution Across Targets

> Histogram of the selected `lambda.1se` across all targets (from the last
> outer fold as representative). A unimodal distribution means the data has
> a consistent signal strength. Bimodality would suggest two populations of
> targets — some with strong regulatory signal (low λ) and others that are
> essentially null (high λ, near-intercept-only model).

```{r lambda-dist, fig.height=6}
lam_df <- bind_rows(
  tibble(ll = log(lambda_selected_A[[outer_k]]), model = "A: residuals ~ proteins"),
  tibble(ll = log(lambda_selected_B[[outer_k]]), model = "B: residuals ~ mRNAs")
) %>% filter(is.finite(ll))

ggplot(lam_df, aes(x = ll, fill = model)) +
  geom_histogram(bins = 50, alpha = 0.5, colour = "black",
                 linewidth = 0.2, position = "identity") +
  scale_fill_manual(values = model_colours[1:2]) +
  labs(x = "log(λ.1se)", y = "# targets", fill = "Model",
       title = "Distribution of Selected Lambda Across Targets",
       subtitle = "Unimodal = consistent signal; bimodal = two regimes") +
  theme_report()
```


# Model Performance {.tabset}

> **Aim:** Compare the three models' predictive performance on protein
> residuals. R² is the primary metric because it is scale-free and directly
> interpretable as the proportion of post-transcriptional variance explained.
> The key comparisons are: (1) A vs C — does the model beat random?
> (2) A vs B — do protein predictors outperform mRNA predictors for
> explaining residuals? If B performs nearly as well as A, the regulators
> may reflect co-transcriptional leakage rather than genuine post-
> transcriptional biology.

## R² Summary Table

```{r r2-table}
r2_df %>%
  group_by(model) %>%
  summarise(n = n(), median = median(r2), mean = mean(r2),
            pct_pos = mean(r2 > 0) * 100, pct_10 = mean(r2 > 0.1) * 100,
            pct_30 = mean(r2 > 0.3) * 100, .groups = "drop") %>%
  kable(digits = 3,
        col.names = c("Model", "N", "Median R²", "Mean R²",
                       "% > 0", "% > 0.1", "% > 0.3"),
        caption = "R² summary (out-of-sample, 10-fold CV on protein residuals)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## R² Histograms

```{r r2-hist, fig.height=6}
ggplot(r2_df, aes(x = r2, fill = model)) +
  geom_histogram(position = "identity", bins = 80,
                 alpha = 0.35, colour = "black", linewidth = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
  scale_fill_manual(values = model_colours) +
  coord_cartesian(xlim = c(-0.5, 1)) +
  labs(x = "R² (out-of-sample)", y = "# targets", fill = "Model",
       title = "R² Distributions — Residual Prediction",
       subtitle = "All R² values are purely post-transcriptional (Step 0 residuals)") +
  theme_report()
```

```{r r2-density, fig.height=5}
ggplot(r2_df, aes(x = r2, colour = model, fill = model)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_colour_manual(values = model_colours) +
  scale_fill_manual(values = model_colours) +
  coord_cartesian(xlim = c(-0.5, 1)) +
  labs(x = "R²", y = "Density", colour = "Model", fill = "Model",
       title = "R² Density Comparison") +
  theme_report()
```

## MSE Distributions

```{r mse-hist, fig.height=5}
mse_long <- bind_rows(
  tibble(target = genes, mse = mean_mse_A, model = "A: residuals ~ proteins"),
  tibble(target = genes, mse = mean_mse_B, model = "B: residuals ~ mRNAs"),
  tibble(target = genes, mse = mean_mse_C, model = "C: permuted baseline")
) %>% filter(is.finite(mse)) %>%
  mutate(model = factor(model, levels = names(model_colours)))

cap <- as.numeric(quantile(mse_long$mse, 0.999))
ggplot(mse_long %>% filter(mse <= cap), aes(x = mse, fill = model)) +
  geom_histogram(position = "identity", bins = 100,
                 alpha = 0.35, colour = "black", linewidth = 0.2) +
  scale_fill_manual(values = model_colours) +
  labs(x = "Mean MSE (10-fold CV)", y = "# targets", fill = "Model",
       title = "MSE Distributions") +
  theme_report()
```

## Model A vs B: Per-Target Scatter

> **What this answers:** For each target, is it better predicted by proteins
> (Model A) or mRNAs (Model B)? Points above the diagonal mean Model A wins.
> The percentage of targets where A > B quantifies the overall advantage of
> protein-level predictors for explaining post-transcriptional residuals.

```{r a-vs-b, fig.height=7}
ab_df <- tibble(target = genes, r2_A = r2_A, r2_B = r2_B) %>%
  filter(is.finite(r2_A) & is.finite(r2_B)) %>%
  mutate(delta = r2_A - r2_B, a_wins = delta > 0)

pct_a_wins <- mean(ab_df$a_wins, na.rm = TRUE) * 100
top_a <- ab_df %>% slice_max(delta, n = 10, with_ties = FALSE)

ggplot(ab_df, aes(x = r2_B, y = r2_A)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(aes(colour = a_wins), alpha = 0.4, size = 1) +
  geom_text_repel(data = top_a, aes(label = target),
                  size = 2.8, max.overlaps = 15, colour = "red3") +
  scale_colour_manual(values = c("TRUE" = "steelblue", "FALSE" = "grey70"),
                      labels = c("TRUE" = "A > B", "FALSE" = "B ≥ A")) +
  coord_cartesian(xlim = c(-0.5, 1), ylim = c(-0.5, 1)) +
  labs(x = "R² — Model B (residuals ~ mRNAs)",
       y = "R² — Model A (residuals ~ proteins)",
       colour = "Better model",
       title = "Per-Target R²: Model A vs B",
       subtitle = sprintf("%.1f%% of targets better predicted by proteins", pct_a_wins)) +
  theme_report()
```

## Variance Decomposition

> **What this answers:** Across all targets, what fraction show strong protein
> signal, moderate improvement, mRNA-dominant signal (co-transcriptional
> leakage), or no signal at all? This is a high-level summary of where
> post-transcriptional regulation matters.

```{r decomp, fig.height=4}
decomp <- ab_df %>%
  mutate(cat = case_when(
    r2_A > 0.3 & delta > 0.05 ~ "Strong protein signal (R²>0.3, Δ>0.05)",
    r2_A > 0.1 & delta > 0    ~ "Moderate protein improvement",
    r2_B > r2_A               ~ "mRNA better (co-transcriptional?)",
    TRUE                       ~ "Weak / no signal"
  )) %>% dplyr::count(cat, name = "n") %>% mutate(pct = n / sum(n) * 100)

ggplot(decomp, aes(x = reorder(cat, n), y = n, fill = cat)) +
  geom_col(width = 0.7, alpha = 0.85) +
  geom_text(aes(label = sprintf("%d (%.1f%%)", n, pct)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  labs(x = NULL, y = "# targets", title = "Variance Decomposition") +
  theme_report() + theme(legend.position = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.3)))
```


# Elastic Net Sensitivity {.tabset}

> **Aim:** Test whether the top LASSO regulators are robust to the penalty
> choice. LASSO (α=1) arbitrarily picks one of several correlated predictors
> and zeros out the rest. Elastic net (α=0.5) mixes L1 and L2 penalties,
> retaining correlated predictors together. If the same proteins emerge under
> both penalties, confidence is high. If they diverge, the LASSO regulators
> may be arbitrary selections among correlated candidates (e.g., proteins in
> the same complex).

```{r elastic-net, cache=TRUE}
cat("Running elastic net sensitivity (α = 0.5) on Model A...\n")
enet_sel_counts <- vector("list", outer_k)

for (f in seq_len(outer_k)) {
  cat("  Fold", f, "\n")
  train_idx <- which(fold_id != f)

  X_tr_raw <- t(protein_mat[, train_idx, drop = FALSE])
  X_tr_imp <- impute_with_train_means(X_tr_raw, X_tr_raw)
  X_tr_sc  <- scale_with_train(X_tr_imp, X_tr_imp)
  storage.mode(X_tr_sc) <- "double"
  vx <- apply(X_tr_sc, 2, var, na.rm = TRUE)
  keep <- is.finite(vx) & vx > 0
  X_tr_sc <- X_tr_sc[, keep, drop = FALSE]
  pnames <- colnames(X_tr_sc)
  sel_en <- setNames(integer(length(pnames)), pnames)

  for (target in genes) {
    y_tr <- as.numeric(residual_mat[target, train_idx])
    ok <- is.finite(y_tr); if (sum(ok) < min_n_train) next
    cur <- if (target %in% pnames) pnames[pnames != target] else pnames
    if (length(cur) < 2) next

    cv_en <- cv.glmnet(X_tr_sc[ok, cur, drop = FALSE], y_tr[ok], alpha = 0.5,
                       nfolds = inner_k,
                       foldid = sample(rep(seq_len(inner_k), length.out = sum(ok))),
                       standardize = FALSE, family = "gaussian")
    b <- as.matrix(coef(cv_en, s = cv_en$lambda.1se))[-1, , drop = FALSE]
    active <- intersect(rownames(b)[b[, 1] != 0], names(sel_en))
    if (length(active) > 0) sel_en[active] <- sel_en[active] + 1L
  }
  enet_sel_counts[[f]] <- sel_en
}

enet_mat <- agg_sel(enet_sel_counts)
enet_total <- colSums(enet_mat)
cat("Elastic net complete.\n")
```

## LASSO vs Elastic Net Comparison

> **What this shows:** Scatter plot of per-regulator selection frequency
> under LASSO vs elastic net. High Spearman ρ = regulators are robust.
> Points far off the diagonal are penalty-sensitive — interpret with caution.

```{r lasso-vs-enet, fig.height=7}
lasso_total_A <- colSums(sel_mat_A)

compare_regs <- full_join(
  tibble(protein = names(lasso_total_A), lasso = as.integer(lasso_total_A)),
  tibble(protein = names(enet_total),    enet  = as.integer(enet_total)),
  by = "protein"
) %>% mutate(lasso = replace_na(lasso, 0), enet = replace_na(enet, 0))

cor_val <- cor(compare_regs$lasso, compare_regs$enet,
               method = "spearman", use = "complete.obs")

top30 <- compare_regs %>%
  filter(lasso >= sort(lasso, decreasing = TRUE)[min(30, nrow(compare_regs))] |
         enet  >= sort(enet,  decreasing = TRUE)[min(30, nrow(compare_regs))])

ggplot(compare_regs, aes(lasso, enet)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.3, size = 1, colour = "grey60") +
  geom_point(data = top30, colour = "steelblue", size = 2, alpha = 0.7) +
  geom_text_repel(data = top30 %>% slice_max(lasso, n = 15),
                  aes(label = protein), size = 2.5, max.overlaps = 20) +
  labs(x = "# targets (LASSO α=1)", y = "# targets (Elastic Net α=0.5)",
       title = "LASSO vs Elastic Net: Regulator Robustness",
       subtitle = sprintf("Spearman ρ = %.3f", cor_val)) +
  theme_report()
```


# Regulator Analysis {.tabset}

> **Aim:** Characterise the protein regulators identified by Model A. A
> "regulator" is a protein selected by the LASSO as a non-zero predictor
> of another protein's residual. We aggregate selection counts across folds,
> compute directionality (positive vs negative), and apply high-confidence
> filters to identify the most robust candidates.

## High-Confidence Regulators

> **Filtering criteria:** A high-confidence (HC) regulator must be:
> (1) selected in ≥ 8 of 10 outer folds (reproducible across data splits),
> (2) selected for ≥ 50 distinct targets (broad influence),
> (3) directionally stable (|frac_pos − frac_neg| ≥ 0.5, consistent sign).

```{r hc-regs-A}
dir_A <- coef_A_df %>%
  mutate(sn = sign(beta)) %>%
  group_by(regulator) %>%
  summarise(n = n(), fp = mean(sn > 0), fn = mean(sn < 0),
            dom = if_else(fp >= fn, "positive", "negative"),
            str = abs(fp - fn), .groups = "drop")

lasso_folds_A <- colSums(sel_mat_A > 0)

reg_tbl_A <- tibble(protein = names(lasso_total_A),
                    total = as.integer(lasso_total_A),
                    folds = as.integer(lasso_folds_A)) %>%
  left_join(dir_A, by = c("protein" = "regulator")) %>%
  mutate(across(c(fp, fn, str), ~ replace_na(.x, 0)))

hc_A <- reg_tbl_A %>%
  filter(folds >= hc_min_folds, total >= hc_min_targets, str >= hc_min_strength) %>%
  arrange(desc(total))

cat(sprintf("HC regulators (Model A): %d (%d pos, %d neg)\n",
            nrow(hc_A), sum(hc_A$dom == "positive"), sum(hc_A$dom == "negative")))
```

```{r hc-top15, fig.height=7}
top_hc <- hc_A %>%
  group_by(dom) %>% slice_max(total, n = 15, with_ties = FALSE) %>% ungroup()

ggplot(top_hc, aes(x = reorder(protein, total), y = total, fill = dom)) +
  geom_col(width = 0.8) + coord_flip() +
  facet_wrap(~ dom, scales = "free_y") +
  scale_fill_manual(values = sign_colours) +
  labs(x = "Regulator", y = "# targets selected",
       title = "Top 15 HC Regulators (Model A)") +
  theme_report() + theme(legend.position = "none", strip.background = element_blank())
```

## Directionality Scatter

> **What this shows:** Each dot is a regulator protein. X = breadth (how many
> targets it predicts), Y = directionality (how consistently positive or
> negative). The green box marks the HC region. Regulators inside this box
> are reproducible, broad, and directionally consistent.

```{r dir-scatter, fig.height=6}
ggplot(reg_tbl_A %>% filter(dom %in% c("positive", "negative")) %>%
         mutate(pass = folds >= hc_min_folds),
       aes(total, str)) +
  annotate("rect", xmin = hc_min_targets, xmax = Inf,
           ymin = hc_min_strength, ymax = Inf,
           fill = "darkseagreen2", alpha = 0.2) +
  geom_point(aes(colour = dom, alpha = pass), size = 0.8) +
  scale_alpha_manual(values = c(`TRUE` = 0.7, `FALSE` = 0.1), guide = "none") +
  geom_vline(xintercept = hc_min_targets, linetype = "dashed") +
  geom_hline(yintercept = hc_min_strength, linetype = "dashed") +
  scale_colour_manual(values = sign_colours) +
  labs(x = "# targets selected", y = "|pos − neg|", colour = "Dominant sign",
       title = "Directionality vs Breadth — Model A") +
  theme_report()
```

## Network Breadth

> **What this shows:** How many targets each regulator predicts (left) and
> how many regulators each target has (right). Hub regulators with very high
> breadth are candidate master regulators.

```{r breadth, fig.height=8}
p1 <- coef_A_df %>% distinct(regulator, target) %>% dplyr::count(regulator, name = "n") %>%
  ggplot(aes(n)) + geom_histogram(bins = 50, fill = "darkcyan", colour = "black",
                                   alpha = 0.6, linewidth = 0.2) +
  labs(x = "# targets per regulator", y = "# regulators", title = "Regulator Breadth") +
  theme_report()

p2 <- coef_A_df %>% distinct(target, regulator) %>% dplyr::count(target, name = "n") %>%
  ggplot(aes(n)) + geom_histogram(bins = 50, fill = "darkcyan", colour = "black",
                                   alpha = 0.6, linewidth = 0.2) +
  labs(x = "# regulators per target", y = "# targets", title = "Target Complexity") +
  theme_report()

p1 / p2
```

## Regulators Ranked by R²

> **What this answers:** Which HC regulators are associated with the
> best-predicted targets? A regulator whose targets have high R² is both
> reproducibly selected and contributes to well-predicted outcomes.

```{r reg-r2, fig.height=7}
r2_tgt <- tibble(target = genes, r2 = r2_A) %>% filter(is.finite(r2))

reg_r2 <- coef_A_df %>%
  filter(regulator %in% hc_A$protein) %>%
  distinct(regulator, target) %>%
  left_join(r2_tgt, by = "target") %>% filter(is.finite(r2)) %>%
  group_by(regulator) %>%
  summarise(n = n(), med_r2 = median(r2), .groups = "drop") %>%
  left_join(hc_A %>% dplyr::select(protein, dom), by = c("regulator" = "protein")) %>%
  slice_max(med_r2, n = 20, with_ties = FALSE)

ggplot(reg_r2, aes(reorder(regulator, med_r2), med_r2, fill = dom)) +
  geom_col(width = 0.7, alpha = 0.85) + coord_flip() +
  scale_fill_manual(values = sign_colours) +
  labs(x = "Regulator", y = "Median R² of targets", fill = "Sign",
       title = "Top 20 HC Regulators by Median Target R²") +
  theme_report()
```


# Model A vs B: Regulator Overlap {.tabset}

> **Aim:** This is the critical test for co-transcriptional leakage. We
> identify HC regulators in Model B (mRNAs), then classify all HC regulators
> into three categories:
>
> - **Model A only:** Protein predicts residuals but the corresponding mRNA
>   does not → **genuine post-transcriptional regulator**
> - **Both A and B:** Both protein and mRNA predict residuals → **suspicious,
>   likely co-transcriptional leakage** (the signal survived OLS correction)
> - **Model B only:** mRNA predicts but protein does not → **transcriptional
>   residual** not fully removed by Step 0
>
> The "A only" set represents the highest-confidence post-transcriptional
> candidates and should be the focus of biological follow-up.

## HC Regulators in Model B

```{r hc-B}
lasso_total_B <- colSums(sel_mat_B)
lasso_folds_B <- colSums(sel_mat_B > 0)

dir_B <- coef_B_df %>%
  mutate(sn = sign(beta)) %>%
  group_by(regulator) %>%
  summarise(n = n(), fp = mean(sn > 0), fn = mean(sn < 0),
            dom = if_else(fp >= fn, "positive", "negative"),
            str = abs(fp - fn), .groups = "drop")

reg_tbl_B <- tibble(protein = names(lasso_total_B),
                    total = as.integer(lasso_total_B),
                    folds = as.integer(lasso_folds_B)) %>%
  left_join(dir_B, by = c("protein" = "regulator")) %>%
  mutate(across(c(fp, fn, str), ~ replace_na(.x, 0)))

hc_B <- reg_tbl_B %>%
  filter(folds >= hc_min_folds, total >= hc_min_targets, str >= hc_min_strength) %>%
  arrange(desc(total))

cat(sprintf("HC regulators (Model B — mRNA): %d\n", nrow(hc_B)))
```

## Overlap Classification

```{r overlap}
set_A <- hc_A$protein; set_B <- hc_B$protein
in_both <- intersect(set_A, set_B)
only_A  <- setdiff(set_A, set_B)
only_B  <- setdiff(set_B, set_A)

tribble(
  ~Category,                            ~Count,        ~Interpretation,
  "Model A only (protein, not mRNA)",   length(only_A), "Genuine post-transcriptional",
  "Both A and B",                       length(in_both),"Co-transcriptional leakage?",
  "Model B only (mRNA, not protein)",   length(only_B), "Transcriptional residual"
) %>%
  kable(caption = "Overlap of HC regulators between Models A and B") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r overlap-bar, fig.height=5}
bind_rows(
  tibble(gene = only_A, cat = "A only\n(post-transcriptional)"),
  tibble(gene = in_both, cat = "Both A & B\n(leakage?)"),
  tibble(gene = only_B, cat = "B only\n(transcriptional)")
) %>% dplyr::count(cat, name = "n") %>%
  mutate(cat = factor(cat, levels = c("A only\n(post-transcriptional)",
                                       "Both A & B\n(leakage?)",
                                       "B only\n(transcriptional)"))) %>%
  ggplot(aes(cat, n, fill = cat)) +
  geom_col(width = 0.6, alpha = 0.85, colour = "black", linewidth = 0.2) +
  geom_text(aes(label = n), vjust = -0.5, size = 4.5) +
  scale_fill_manual(values = c("#E64B35", "#91D1C2", "#4DBBD5")) +
  labs(x = NULL, y = "# HC regulators",
       title = "Model A vs B Overlap: The Co-Transcriptional Leakage Test",
       subtitle = "'A only' = highest-confidence post-transcriptional candidates") +
  theme_report() + theme(legend.position = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))
```

## Leakage Candidates (Both A & B)

> **What to do with these:** These genes' protein AND mRNA levels both predict
> targets' residuals. Check whether they are known co-transcriptionally
> regulated clusters (e.g., mitochondrial biogenesis genes under PGC-1α). If
> the sign is the same in both models, co-transcriptional leakage is very
> likely. If signs differ, the protein may have genuine post-transcriptional
> activity in addition to its transcriptional co-regulation.

```{r leakage-table}
if (length(in_both) > 0) {
  hc_A %>%
    filter(protein %in% in_both) %>%
    dplyr::select(protein, total_A = total, sign_A = dom) %>%
    left_join(hc_B %>% dplyr::select(protein, total_B = total, sign_B = dom),
              by = "protein") %>%
    mutate(same_sign = sign_A == sign_B) %>% arrange(desc(total_A)) %>%
    kable(caption = "Genes in both Model A and B HC lists") %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
} else {
  cat("✓ No overlap — all Model A HC regulators are unique to the protein level.\n")
}
```

## Highest-Confidence Post-Transcriptional Regulators (A only)

> **These are the final output.** These proteins pass all statistical filters
> AND cannot be recapitulated at the mRNA level, making them the strongest
> candidates for genuine post-transcriptional regulation.

```{r a-only-table}
hc_A %>%
  filter(protein %in% only_A) %>%
  dplyr::select(protein, total, folds, dom, str) %>%
  kable(col.names = c("Protein", "# targets", "# folds", "Sign", "Directionality"),
        digits = 3,
        caption = "Highest-confidence post-transcriptional regulators (A only)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```


# External Validation {.tabset}

> **Aim:** Models trained on Satpathy LSCC tumour are evaluated on independent
> datasets from different studies, tissues, and conditions. Generalisation to
> external data is the gold standard for model validity — a model that performs
> well only on training data is overfit. Here we:
>
> 1. Retrain Model A on **all** Satpathy samples (no hold-out) to get final
>    per-target models.
> 2. Apply training-set OLS coefficients (Step 0) to compute residuals in each
>    external dataset — this prevents data leakage.
> 3. Predict external residuals using the final Model A and compute R²/MSE.
> 4. Repeat with Model C (permuted) as a null baseline.

## Retrain Final Models on Full Satpathy

> **What this does:** The CV loop used partial training sets (9/10 folds).
> For external validation, we retrain on all Satpathy samples to get the
> best possible models. The OLS intercepts, slopes, and protein scaling
> statistics are saved as a bundle for applying to external data.

```{r final-train, cache=TRUE}
cat("Retraining final models on all Satpathy samples...\n")

# ── Full protein predictor matrix ──
X_full_raw <- t(protein_mat)  # samples × genes
X_full_imp <- impute_with_train_means(X_full_raw, X_full_raw)
X_full_sc  <- scale_with_train(X_full_imp, X_full_imp)
storage.mode(X_full_sc) <- "double"

vx_full <- apply(X_full_sc, 2, var, na.rm = TRUE)
keep_full <- is.finite(vx_full) & vx_full > 0
X_full_sc <- X_full_sc[, keep_full, drop = FALSE]
final_pred_names <- colnames(X_full_sc)

# Save scaling stats for external application
prot_mu_full <- colMeans(X_full_imp, na.rm = TRUE)
prot_sd_full <- apply(X_full_imp, 2, sd, na.rm = TRUE)
prot_sd_full[!is.finite(prot_sd_full) | prot_sd_full == 0] <- 1

# ── Fit final LASSO per target ──
final_models_A <- vector("list", n_genes); names(final_models_A) <- genes
final_models_C <- vector("list", n_genes); names(final_models_C) <- genes
final_lambda_A <- setNames(rep(NA_real_, n_genes), genes)
final_lambda_C <- setNames(rep(NA_real_, n_genes), genes)
final_feat_A   <- vector("list", n_genes); names(final_feat_A) <- genes

n_fit <- 0L
for (ii in seq_along(genes)) {
  target <- genes[ii]
  if (ii %% 500 == 0) cat("  Progress:", ii, "/", n_genes, "\n")

  y_tr <- as.numeric(residual_mat[target, ])
  ok <- is.finite(y_tr); if (sum(ok) < min_n_train) next

  cur <- if (target %in% final_pred_names) {
    final_pred_names[final_pred_names != target]
  } else final_pred_names
  if (length(cur) < 2) next

  Xtr <- X_full_sc[ok, cur, drop = FALSE]
  fid <- sample(rep(seq_len(inner_k), length.out = sum(ok)))

  cv_A <- cv.glmnet(Xtr, y_tr[ok], alpha = 1, nfolds = inner_k, foldid = fid,
                    standardize = FALSE, family = "gaussian")
  final_models_A[[target]] <- cv_A
  final_lambda_A[target]   <- cv_A$lambda.1se
  final_feat_A[[target]]   <- cur

  y_perm <- sample(y_tr[ok], replace = FALSE)
  cv_C <- cv.glmnet(Xtr, y_perm, alpha = 1, nfolds = inner_k, foldid = fid,
                    standardize = FALSE, family = "gaussian")
  final_models_C[[target]] <- cv_C
  final_lambda_C[target]   <- cv_C$lambda.1se

  n_fit <- n_fit + 1L
}
cat("Final models trained:", n_fit, "targets\n")
```

## Evaluate on External Datasets

> **What this does:** For each external dataset, we: (1) align genes/samples
> with the training set, (2) compute protein residuals using **training-set**
> OLS coefficients (no data leakage), (3) build a scaled protein predictor
> matrix using training-set means/SDs, (4) predict residuals using the final
> Model A and C, and (5) compute MSE and R² per target.

```{r external-eval, cache=TRUE}
# Helper: build scaled predictor matrix for external data
make_X_ext <- function(prot_test, pred_names, mu, sdv) {
  X_raw <- t(prot_test)
  genes_present <- colnames(X_raw)
  mu_p <- mu[genes_present]; mu_p[!is.finite(mu_p)] <- 0
  sd_p <- sdv[genes_present]; sd_p[!is.finite(sd_p) | sd_p == 0] <- 1

  for (j in seq_len(ncol(X_raw))) {
    nas <- is.na(X_raw[, j]) | !is.finite(X_raw[, j])
    if (any(nas)) X_raw[nas, j] <- mu_p[j]
  }
  X_sc <- sweep(sweep(X_raw, 2, mu_p, "-"), 2, sd_p, "/")
  storage.mode(X_sc) <- "double"

  X_full <- matrix(0, nrow = nrow(X_sc), ncol = length(pred_names),
                   dimnames = list(rownames(X_sc), pred_names))
  common <- intersect(pred_names, colnames(X_sc))
  if (length(common) > 0) X_full[, common] <- X_sc[, common, drop = FALSE]
  storage.mode(X_full) <- "double"
  X_full
}

ext_results <- list()
ext_qc_rows <- list()

for (i in seq_len(nrow(ext_test_sets))) {
  ds   <- ext_test_sets$dataset[i]
  cat("\n--- Evaluating:", ds, "---\n")

  if (!file.exists(ext_test_sets$protein_path[i]) ||
      !file.exists(ext_test_sets$mrna_path[i])) {
    cat("  Files not found, skipping.\n")
    ext_qc_rows[[i]] <- tibble(dataset = ds, n_samples = NA, n_genes = 0, n_eval = 0)
    next
  }

#  prot_ext <- readRDS(ext_test_sets$protein_path[i])
 # mrna_ext <- readRDS(ext_test_sets$mrna_path[i])
  
  
  prot_ext_raw <- read_csv(ext_test_sets$protein_path[i], show_col_types = FALSE)
  prot_ext <- prot_ext_raw %>% dplyr::select(where(is.numeric)) %>% as.matrix()
  rownames(prot_ext) <- prot_ext_raw$symbol

  mrna_ext_raw <- read_csv(ext_test_sets$mrna_path[i], show_col_types = FALSE)
  mrna_ext <- mrna_ext_raw %>% dplyr::select(where(is.numeric)) %>% as.matrix()
  rownames(mrna_ext) <- mrna_ext_raw$symbol

  # Align samples
  cs <- intersect(colnames(prot_ext), colnames(mrna_ext))
  if (length(cs) == 0) { cat("  No common samples.\n"); next }
  prot_ext <- prot_ext[, cs, drop = FALSE]
  mrna_ext <- mrna_ext[, cs, drop = FALSE]

  # Align genes with training
  cg <- intersect(genes, intersect(rownames(prot_ext), rownames(mrna_ext)))
  if (length(cg) == 0) { cat("  No common genes.\n"); next }
  prot_ext <- prot_ext[cg, , drop = FALSE]
  mrna_ext <- mrna_ext[cg, , drop = FALSE]
  cat("  Samples:", ncol(prot_ext), " Genes:", length(cg), "\n")

  # ── Compute residuals using TRAINING-set OLS coefficients ──
  resid_ext <- matrix(NA_real_, nrow = length(cg), ncol = ncol(prot_ext),
                      dimnames = list(cg, colnames(prot_ext)))
  for (g in cg) {
    if (!is.finite(ols_slope[g]) || !is.finite(ols_intercept[g])) next
    p <- as.numeric(prot_ext[g, ])
    m <- as.numeric(mrna_ext[g, ])
    ok <- is.finite(p) & is.finite(m)
    resid_ext[g, which(ok)] <- p[ok] - (ols_intercept[g] + ols_slope[g] * m[ok])
  }

  # ── Build scaled protein predictor matrix ──
  X_ext <- make_X_ext(prot_ext, final_pred_names, prot_mu_full, prot_sd_full)

  # ── Predictor coverage QC ──
  # For each target with a trained model, check what fraction of its
  # non-zero LASSO coefficients correspond to proteins present in
  # this validation dataset.
  ext_proteins_available <- rownames(prot_ext)

  coverage_per_target <- vapply(cg, function(target) {
    if (is.null(final_models_A[[target]])) return(NA_real_)
    if (!is.finite(final_lambda_A[target])) return(NA_real_)

    b <- as.matrix(coef(final_models_A[[target]], s = final_lambda_A[target]))[-1, , drop = FALSE]
    active <- rownames(b)[b[, 1] != 0]
    if (length(active) == 0) return(NA_real_)

    mean(active %in% ext_proteins_available) * 100
  }, numeric(1))

  coverage_per_target <- coverage_per_target[is.finite(coverage_per_target)]

  cat(sprintf("  Predictor coverage: median = %.1f%%, mean = %.1f%%, min = %.1f%%\n",
              median(coverage_per_target),
              mean(coverage_per_target),
              min(coverage_per_target)))

  # ── Add to QC row ──
  # Replace your existing ext_qc_rows[[i]] line with this:
  ext_qc_rows[[i]] <- tibble(
    dataset    = ds,
    n_samples  = ncol(prot_ext),
    n_genes    = length(cg),
    n_eval     = n_eval,
    med_coverage_pct = median(coverage_per_target),
    mean_coverage_pct = mean(coverage_per_target),
    min_coverage_pct  = min(coverage_per_target)
  )
  
  
  # ── Predict per target ──
  mse_ext_A <- mse_ext_C <- setNames(rep(NA_real_, length(cg)), cg)

  for (target in cg) {
    if (is.null(final_models_A[[target]])) next
    if (!is.finite(final_lambda_A[target])) next
    feat <- final_feat_A[[target]]; if (is.null(feat)) next

    y_te <- as.numeric(resid_ext[target, ])
    ok_te <- is.finite(y_te); if (sum(ok_te) < 2) next

    Xte <- X_ext[ok_te, feat, drop = FALSE]

    yhat_A <- as.numeric(predict(final_models_A[[target]], newx = Xte,
                                  s = final_lambda_A[target]))
    mse_ext_A[target] <- mean((y_te[ok_te] - yhat_A)^2)

    yhat_C <- as.numeric(predict(final_models_C[[target]], newx = Xte,
                                  s = final_lambda_C[target]))
    mse_ext_C[target] <- mean((y_te[ok_te] - yhat_C)^2)
  }

  n_eval <- sum(is.finite(mse_ext_A))
  cat("  Evaluated targets:", n_eval, "\n")

  ext_results[[ds]] <- tibble(
    target = names(mse_ext_A), mse_real = mse_ext_A, mse_rand = mse_ext_C,
    dataset = ds
  ) %>% filter(is.finite(mse_real) & is.finite(mse_rand))

  ext_qc_rows[[i]] <- tibble(dataset = ds, n_samples = ncol(prot_ext),
                              n_genes = length(cg), n_eval = n_eval)
}

ext_mse_long <- bind_rows(ext_results)
ext_qc_tbl   <- bind_rows(ext_qc_rows)
has_ext <- nrow(ext_mse_long) > 0
```

## External QC Table

> **What this shows:** How many samples and genes overlap between each
> external dataset and the Satpathy training set. Low overlap reduces power
> but does not invalidate the test.

```{r ext-qc, eval=has_ext}
ext_qc_tbl %>%
  kable(col.names = c("Dataset", "Samples", "Overlapping genes", "Evaluated targets",
              "Median coverage %", "Mean coverage %", "Min coverage %")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)


```

## External R² Distributions

> **What this answers:** Does the model generalise? R² is computed using
> training-set residual variance as denominator, making it comparable across
> datasets. Positive external R² = the model captures genuine biology.
> Consistent with training CV = not overfit. Large leftward shift = overfit.

```{r ext-r2, eval=has_ext, fig.height=8}
ext_r2 <- ext_mse_long %>%
  mutate(var_tr = resid_var[target],
         r2_real = 1 - mse_real / var_tr,
         r2_rand = 1 - mse_rand / var_tr) %>%
  filter(is.finite(r2_real) & is.finite(r2_rand))

ext_r2_long <- ext_r2 %>%
  pivot_longer(c(r2_real, r2_rand), names_to = "model", values_to = "r2") %>%
  mutate(model = recode(model, r2_real = "Real (A)", r2_rand = "Random (C)"))

ext_r2_long %>%
  group_by(dataset) %>%
  group_map(function(df, key) {
    ggplot(df, aes(r2, fill = model)) +
      geom_histogram(position = "identity", bins = 60,
                     alpha = 0.4, colour = "black", linewidth = 0.15) +
      geom_vline(xintercept = 0, linetype = "dashed") +
      coord_cartesian(xlim = c(-1, 1)) +
      labs(x = "R²", y = "# targets", title = key$dataset, fill = NULL) +
      theme_report(10) + theme(legend.position = "bottom")
  }, .keep = TRUE) %>%
  wrap_plots(ncol = 2) +
  plot_annotation(title = "External Validation: R² Distributions",
                  subtitle = "Real model should shift right of random baseline")
```

## External R² Summary

```{r ext-r2-summary, eval=has_ext}
ext_r2 %>%
  group_by(dataset) %>%
  summarise(n = n(), med_r2 = median(r2_real), mean_r2 = mean(r2_real),
            pct_pos = mean(r2_real > 0) * 100,
            med_r2_rand = median(r2_rand), .groups = "drop") %>%
  kable(digits = 3,
        col.names = c("Dataset", "N", "Median R²", "Mean R²", "% > 0",
                       "Median R² (random)"),
        caption = "External validation R² summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Combined R²: Training CV + External

> **What this answers:** Are the training and external R² distributions
> consistent? Overlay allows direct visual comparison. A large leftward
> shift in external R² relative to training CV indicates overfitting.

```{r combined-r2, eval=has_ext, fig.height=5}
r2_combined <- bind_rows(
  r2_df %>% filter(model == "A: residuals ~ proteins") %>%
    mutate(source = "Training CV (10-fold)") %>% dplyr::select(target, r2, source),
  ext_r2 %>% mutate(source = paste0("Ext: ", dataset)) %>%
    dplyr::select(target, r2 = r2_real, source)
)

ggplot(r2_combined, aes(r2, fill = source)) +
  geom_histogram(aes(y = after_stat(density)),
                 position = "identity", bins = 60,
                 alpha = 0.25, colour = "black", linewidth = 0.15) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(-1, 1)) +
  labs(x = "R²", y = "Density", fill = "Source",
       title = "R² Comparison: Training CV vs External Validation",
       subtitle = "Consistent distributions = model generalises well") +
  theme_report() + theme(legend.position = "bottom")
```


# GO Enrichment {.tabset}

> **Aim:** Test whether HC regulators enrich for expected biological processes.
> This provides face validity:
>
> - **Positive regulators** should enrich for translation, ribosome biogenesis,
>   mRNA processing — processes that increase protein production.
> - **Negative regulators** should enrich for protein degradation, ubiquitin-
>   proteasome, autophagy, stress response — processes that decrease protein.
> - **"A only"** regulators should show the cleanest enrichment for genuinely
>   post-transcriptional terms, with minimal leakage of transcriptional
>   processes like mitochondrial biogenesis.

```{r go-helper}
run_topgo <- function(fg, bg, ontology = "BP", n_top = 20) {
  glist <- factor(as.integer(bg %in% fg)); names(glist) <- bg
  GOdata <- new("topGOdata", ontology = ontology, allGenes = glist,
                geneSel = function(x) x == 1, nodeSize = 10,
                annot = annFUN.org, mapping = "org.Hs.eg.db", ID = "symbol")
  result <- runTest(GOdata, algorithm = "weight01", statistic = "fisher")
  GenTable(GOdata, Fisher = result, orderBy = "Fisher",
           ranksOf = "Fisher", topNodes = n_top)
}
bg_genes <- rownames(protein_mat)
```

## Positive HC Regulators

```{r go-pos, fig.height=6}
fg_pos <- hc_A %>% filter(dom == "positive") %>% pull(protein)
go_pos <- run_topgo(fg_pos, bg_genes)

go_pos %>% kable(caption = "GO BP — Positive HC regulators") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, font_size = 11)

go_pos %>%
  mutate(Fisher = as.numeric(gsub("< ", "", Fisher)), nlp = -log10(Fisher)) %>%
  slice_max(nlp, n = 15) %>%
  ggplot(aes(nlp, reorder(Term, nlp))) +
  geom_col(fill = "firebrick", alpha = 0.7, width = 0.7) +
  labs(x = expression(-log[10](p)), y = NULL,
       title = "GO: Positive HC Regulators",
       subtitle = "Expected: translation, ribosome biogenesis") +
  theme_report()
```

## Negative HC Regulators

```{r go-neg, fig.height=6}
fg_neg <- hc_A %>% filter(dom == "negative") %>% pull(protein)
go_neg <- run_topgo(fg_neg, bg_genes)

go_neg %>% kable(caption = "GO BP — Negative HC regulators") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, font_size = 11)

go_neg %>%
  mutate(Fisher = as.numeric(gsub("< ", "", Fisher)), nlp = -log10(Fisher)) %>%
  slice_max(nlp, n = 15) %>%
  ggplot(aes(nlp, reorder(Term, nlp))) +
  geom_col(fill = "deepskyblue", alpha = 0.7, width = 0.7) +
  labs(x = expression(-log[10](p)), y = NULL,
       title = "GO: Negative HC Regulators",
       subtitle = "Expected: degradation, ubiquitin, stress") +
  theme_report()
```

## "A only" Regulators

```{r go-aonly, fig.height=6}
if (length(only_A) >= 5) {
  go_aonly <- run_topgo(only_A, bg_genes)
  go_aonly %>% kable(caption = "GO BP — 'A only' post-transcriptional regulators") %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, font_size = 11)
  go_aonly %>%
    mutate(Fisher = as.numeric(gsub("< ", "", Fisher)), nlp = -log10(Fisher)) %>%
    slice_max(nlp, n = 15) %>%
    ggplot(aes(nlp, reorder(Term, nlp))) +
    geom_col(fill = "#E64B35", alpha = 0.7, width = 0.7) +
    labs(x = expression(-log[10](p)), y = NULL,
         title = "GO: 'A only' Post-Transcriptional Regulators") +
    theme_report()
} else {
  cat("Too few 'A only' regulators for GO.\n")
}
```

## mTOR / ERK / 5'TOP Analysis {.tabset}

> **Aim:** Test whether the identified regulators show specific relationships
> with known mTOR and ERK pathway markers, and whether 5'TOP genes (whose
> translation is directly controlled by mTOR via 4E-BP1/eIF4E) are
> preferentially explained by Model A.
>
> **Marker definitions:**
>
> - **Positive markers** (mTOR/ERK): expression decreases upon pathway
>   inhibition → they track with pathway activity.
> - **Negative markers** (mTOR/ERK): expression increases upon pathway
>   inhibition → they are inversely related to pathway activity.
> - **5'TOP genes:** contain a 5' terminal oligopyrimidine tract; their
>   translation is directly regulated by mTOR–4E-BP1–eIF4E signalling.
### Load Gene Lists

```{r load-gene-lists}
top_genes <- trimws(readLines(top_genes_path))
top_genes <- top_genes[nzchar(top_genes)]

mtor_pos <- trimws(readLines(mtor_pos_path))
mtor_pos <- mtor_pos[nzchar(mtor_pos)]

mtor_neg <- trimws(readLines(mtor_neg_path))
mtor_neg <- mtor_neg[nzchar(mtor_neg)]

erk_pos <- trimws(readLines(erk_pos_path))
erk_pos <- erk_pos[nzchar(erk_pos)]

erk_neg <- trimws(readLines(erk_neg_path))
erk_neg <- erk_neg[nzchar(erk_neg)]

mtor_all <- union(mtor_pos, mtor_neg)
erk_all  <- union(erk_pos, erk_neg)

cat(sprintf("5'TOP genes: %d (%d in filtered data)\n",
            length(top_genes), sum(top_genes %in% genes)))
cat(sprintf("mTOR markers: %d pos, %d neg (%d total in filtered data)\n",
            length(mtor_pos), length(mtor_neg), sum(mtor_all %in% genes)))
cat(sprintf("ERK markers:  %d pos, %d neg (%d total in filtered data)\n",
            length(erk_pos), length(erk_neg), sum(erk_all %in% genes)))
```

### 5'TOP Genes as Targets: Are They Better Predicted by Model A?

> **What this answers:** If mTOR-driven translational control is a major
> post-transcriptional mechanism, then 5'TOP genes — whose translation is
> directly regulated by mTOR — should have higher Model A R² than non-5'TOP
> genes. Their protein residuals should be more predictable by other proteins
> because the translational machinery (ribosomal proteins, eIF factors) that
> controls them will appear as LASSO regulators.

```{r 5top-targets, fig.height=5}
r2_tgt_df <- tibble(target = genes, r2_A = r2_A, r2_B = r2_B) %>%
  filter(is.finite(r2_A)) %>%
  mutate(is_5top = target %in% top_genes,
         group   = if_else(is_5top, "5'TOP", "Other"))

# ── Boxplot ──
ggplot(r2_tgt_df, aes(x = group, y = r2_A, fill = group)) +
  geom_boxplot(width = 0.5, alpha = 0.7, outlier.alpha = 0.3) +
  geom_jitter(width = 0.15, size = 0.8, alpha = 0.2) +
  scale_fill_manual(values = c("5'TOP" = "#E64B35", "Other" = "grey70")) +
  labs(x = NULL, y = "R² (Model A, out-of-sample)",
       title = "Are 5'TOP Gene Residuals Better Predicted by Proteins?",
       subtitle = sprintf("5'TOP: n=%d, Other: n=%d",
                          sum(r2_tgt_df$is_5top), sum(!r2_tgt_df$is_5top))) +
  theme_report() + theme(legend.position = "none")

# ── Wilcoxon test ──
wt <- wilcox.test(r2_A ~ is_5top, data = r2_tgt_df)
cat(sprintf("Wilcoxon p-value: %.2e\n", wt$p.value))
cat(sprintf("Median R² — 5'TOP: %.3f, Other: %.3f\n",
            median(r2_tgt_df$r2_A[r2_tgt_df$is_5top]),
            median(r2_tgt_df$r2_A[!r2_tgt_df$is_5top])))
```


```{r 5top-a-vs-b, fig.height=5}
# ── Same comparison for Model B — if 5'TOP are also well-predicted by mRNAs,
#    the signal may be co-transcriptional rather than translational ──
r2_tgt_ab <- r2_tgt_df %>% filter(is.finite(r2_B))

ggplot(r2_tgt_ab %>% pivot_longer(c(r2_A, r2_B), names_to = "model", values_to = "r2") %>%
         mutate(model = recode(model, r2_A = "Model A (proteins)", r2_B = "Model B (mRNAs)")),
       aes(x = group, y = r2, fill = model)) +
  geom_boxplot(width = 0.6, alpha = 0.7, position = position_dodge(0.7)) +
  scale_fill_manual(values = c("Model A (proteins)" = "#E64B35",
                                "Model B (mRNAs)" = "#4DBBD5")) +
  labs(x = NULL, y = "R²", fill = NULL,
       title = "5'TOP: Model A vs B",
       subtitle = "If A >> B for 5'TOP → translational; if A ≈ B → co-transcriptional") +
  theme_report()
```

### mTOR/ERK Markers Among HC Regulators

> **What this answers:** Are mTOR or ERK pathway markers over-represented
> among the HC protein regulators? If mTOR-positive markers (proteins that
> track mTOR activity) appear as positive HC regulators, this suggests mTOR
> signalling is a driver of the post-transcriptional signal.

```{r pathway-enrichment-fisher}
run_fisher <- function(fg_regs, pathway_genes, bg, label) {
  a <- sum(fg_regs %in% pathway_genes)
  b <- length(fg_regs) - a
  c <- sum(bg %in% pathway_genes) - a
  d <- length(bg) - a - b - c
  ft <- fisher.test(matrix(c(a, b, c, d), nrow = 2))
  tibble(pathway = label, in_HC = a, not_in_HC = c,
         odds_ratio = ft$estimate, p_value = ft$p.value)
}

hc_proteins <- hc_A$protein

fisher_results <- bind_rows(
  run_fisher(hc_proteins, mtor_pos, bg_genes, "mTOR positive markers"),
  run_fisher(hc_proteins, mtor_neg, bg_genes, "mTOR negative markers"),
  run_fisher(hc_proteins, mtor_all, bg_genes, "mTOR all markers"),
  run_fisher(hc_proteins, erk_pos,  bg_genes, "ERK positive markers"),
  run_fisher(hc_proteins, erk_neg,  bg_genes, "ERK negative markers"),
  run_fisher(hc_proteins, erk_all,  bg_genes, "ERK all markers"),
  run_fisher(hc_proteins, top_genes, bg_genes, "5'TOP genes")
)

fisher_results %>%
  mutate(significant = if_else(p_value < 0.05, "✓", "")) %>%
  kable(digits = c(0, 0, 0, 2, 4, 0),
        col.names = c("Pathway", "In HC regs", "In background only",
                       "Odds ratio", "p (Fisher)", "Sig."),
        caption = "Enrichment of pathway markers among HC regulators") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```


```{r pathway-enrichment-bar, fig.height=5}
fisher_results %>%
  mutate(nlp = -log10(p_value),
         pathway = factor(pathway, levels = rev(pathway))) %>%
  ggplot(aes(x = nlp, y = pathway, fill = nlp > -log10(0.05))) +
  geom_col(width = 0.7, alpha = 0.8) +
  geom_vline(xintercept = -log10(0.05), linetype = "dashed", colour = "red3") +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "grey70"),
                    guide = "none") +
  labs(x = expression(-log[10](p)), y = NULL,
       title = "Pathway Marker Enrichment Among HC Regulators",
       subtitle = "Dashed line = p = 0.05") +
  theme_report()
```

### Sign Concordance: Do mTOR/ERK Markers Match Expected Direction?

> **What this answers:** mTOR-positive markers (expression tracks pathway
> activity) should appear as **positive** HC regulators if they genuinely
> drive translational upregulation. mTOR-negative markers should appear as
> **negative** regulators. If the signs are concordant, the model is
> capturing biologically meaningful directionality. Discordance would suggest
> the model is picking up indirect or confounded associations.

```{r sign-concordance}
hc_with_sign <- hc_A %>% dplyr::select(protein, dom)

sign_check <- function(pathway_genes, pathway_label, expected_sign) {
  hits <- hc_with_sign %>% filter(protein %in% pathway_genes)
  if (nrow(hits) == 0) return(tibble(pathway = pathway_label,
                                      n_hits = 0, concordant = NA, pct = NA))
  tibble(
    pathway    = pathway_label,
    n_hits     = nrow(hits),
    concordant = sum(hits$dom == expected_sign),
    pct        = sum(hits$dom == expected_sign) / nrow(hits) * 100
  )
}

sign_tbl <- bind_rows(
  sign_check(mtor_pos, "mTOR positive → expect positive HC", "positive"),
  sign_check(mtor_neg, "mTOR negative → expect negative HC", "negative"),
  sign_check(erk_pos,  "ERK positive → expect positive HC",  "positive"),
  sign_check(erk_neg,  "ERK negative → expect negative HC",  "negative")
)

sign_tbl %>%
  kable(digits = 1,
        col.names = c("Pathway / expected sign", "N in HC", "Concordant", "% concordant"),
        caption = "Sign concordance: do pathway markers match expected regulator direction?") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Leakage Check: Are mTOR/ERK Markers in "A only" or "Both A & B"?

> **What this answers:** This is the mechanistic crux. If mTOR-positive markers
> appear in "A only" (protein predicts residuals but mRNA doesn't), they are
> acting post-transcriptionally — consistent with mTOR's translational arm
> (4E-BP1/S6K1). If they appear in "Both A & B", the signal may be mTOR's
> transcriptional arm (PGC-1α → mitochondrial gene expression) leaking
> through the OLS correction. This distinction directly addresses the concern
> about mitochondrial protein regulators.


```{r leakage-pathway, fig.height=6}
leakage_class <- tibble(
  gene = c(only_A, in_both, only_B),
  overlap_cat = c(rep("A only", length(only_A)),
                  rep("Both A & B", length(in_both)),
                  rep("B only", length(only_B)))
)

pathway_in_overlap <- leakage_class %>%
  mutate(
    is_mtor_pos = gene %in% mtor_pos,
    is_mtor_neg = gene %in% mtor_neg,
    is_erk_pos  = gene %in% erk_pos,
    is_erk_neg  = gene %in% erk_neg,
    is_5top     = gene %in% top_genes
  )

# ── Summary table ──
leakage_summary <- pathway_in_overlap %>%
  group_by(overlap_cat) %>%
  summarise(
    n_total    = n(),
    mTOR_pos   = sum(is_mtor_pos),
    mTOR_neg   = sum(is_mtor_neg),
    ERK_pos    = sum(is_erk_pos),
    ERK_neg    = sum(is_erk_neg),
    `5TOP`     = sum(is_5top),
    .groups    = "drop"
  )

leakage_summary %>%
  kable(caption = "Pathway markers by overlap category (A only / Both / B only)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# ── Visual ──
leakage_long <- pathway_in_overlap %>%
  pivot_longer(cols = starts_with("is_"),
               names_to = "pathway", values_to = "member") %>%
  filter(member) %>%
  mutate(pathway = recode(pathway,
                          is_mtor_pos = "mTOR+", is_mtor_neg = "mTOR−",
                          is_erk_pos = "ERK+", is_erk_neg = "ERK−",
                          is_5top = "5'TOP"))

if (nrow(leakage_long) > 0) {
  leakage_long %>%
    dplyr::count(overlap_cat, pathway, name = "n") %>%
    mutate(overlap_cat = factor(overlap_cat,
                                levels = c("A only", "Both A & B", "B only"))) %>%
    ggplot(aes(x = pathway, y = n, fill = overlap_cat)) +
    geom_col(position = "dodge", width = 0.7, alpha = 0.85) +
    scale_fill_manual(values = c("A only" = "#E64B35",
                                  "Both A & B" = "#91D1C2",
                                  "B only" = "#4DBBD5")) +
    labs(x = "Pathway marker set", y = "# HC regulators", fill = "Overlap category",
         title = "Where Do Pathway Markers Land?",
         subtitle = "A only = post-transcriptional; Both = co-transcriptional leakage?") +
    theme_report()
} else {
  cat("No pathway markers found among HC regulators.\n")
}
```

### 5'TOP Genes as Targets of mTOR-Positive Regulators

> **What this answers:** The most direct mechanistic prediction: mTOR-positive
> HC regulators (proteins that track mTOR activity) should specifically predict
> 5'TOP gene residuals, because mTOR drives 5'TOP translation via 4E-BP1.
> We test whether the median LASSO coefficient of mTOR-positive regulators
> is higher for 5'TOP targets than for non-5'TOP targets.

```{r 5top-mtor-link, fig.height=5}
# ── Get all Model A coefficients involving mTOR-positive regulators ──
mtor_pos_in_hc <- intersect(mtor_pos, hc_A$protein)

if (length(mtor_pos_in_hc) > 0) {
  mtor_coefs <- coef_A_df %>%
    filter(regulator %in% mtor_pos_in_hc) %>%
    mutate(target_is_5top = target %in% top_genes,
           target_group = if_else(target_is_5top, "5'TOP target", "Other target"))

  if (nrow(mtor_coefs) > 0) {
    # ── Boxplot of absolute beta ──
    ggplot(mtor_coefs, aes(x = target_group, y = abs(beta), fill = target_group)) +
      geom_boxplot(width = 0.5, alpha = 0.7, outlier.alpha = 0.3) +
      scale_fill_manual(values = c("5'TOP target" = "#E64B35", "Other target" = "grey70")) +
      labs(x = NULL, y = "|β| (LASSO coefficient)",
           title = "mTOR-Positive Regulators: Effect Size on 5'TOP vs Other Targets",
           subtitle = sprintf("Regulators tested: %s",
                              paste(mtor_pos_in_hc, collapse = ", "))) +
      theme_report() + theme(legend.position = "none")

    wt2 <- wilcox.test(abs(beta) ~ target_is_5top, data = mtor_coefs)
    cat(sprintf("Wilcoxon p-value (|β| for 5'TOP vs other): %.2e\n", wt2$p.value))
    cat(sprintf("Median |β| — 5'TOP targets: %.4f, Other: %.4f\n",
                median(abs(mtor_coefs$beta[mtor_coefs$target_is_5top])),
                median(abs(mtor_coefs$beta[!mtor_coefs$target_is_5top]))))
  } else {
    cat("No coefficients found for mTOR-positive HC regulators.\n")
  }
} else {
  cat("No mTOR-positive markers among HC regulators — cannot test 5'TOP link.\n")
}
```

# Summary {.tabset}

> **Aim:** Compile all key metrics into a single reference table for quick
> consultation. This table should contain enough information to write the
> results paragraph of a paper.

## Key Metrics

```{r summary-table}
tribble(
  ~Metric,                                     ~Value,
  "Training dataset",                          "Satpathy 2021 LSCC tumour",
  "Filtered genes × samples",                 sprintf("%d × %d", n_genes, n_samples),
  "Median per-gene mRNA–protein ρ",            sprintf("%.3f", med_gene_cor),
  "Median per-sample mRNA–protein ρ",          sprintf("%.3f", med_cor),
  "Outlier samples flagged",                   as.character(nrow(outlier_samples)),
  "Median OLS R² (protein ~ own mRNA)",        sprintf("%.3f", med_ols_r2),
  "HC regulators (Model A — protein)",         as.character(nrow(hc_A)),
  "  — positive",                              as.character(sum(hc_A$dom == "positive")),
  "  — negative",                              as.character(sum(hc_A$dom == "negative")),
  "  — post-transcriptional (A only)",         as.character(length(only_A)),
  "  — co-transcriptional leakage (A∩B)",      as.character(length(in_both)),
  "HC regulators (Model B — mRNA)",            as.character(nrow(hc_B)),
  "Median R² — Model A (CV)",                  sprintf("%.3f", median(r2_A, na.rm = TRUE)),
  "Median R² — Model B (CV)",                  sprintf("%.3f", median(r2_B, na.rm = TRUE)),
  "Median R² — Model C (CV)",                  sprintf("%.3f", median(r2_C, na.rm = TRUE)),
  "% targets A > B",                           sprintf("%.1f%%", pct_a_wins),
  "LASSO vs Elastic Net ρ",                    sprintf("%.3f", cor_val),
  "External datasets evaluated",               as.character(nrow(ext_qc_tbl))
) %>%
  kable(caption = "Key analysis metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Session Info

```{r session-info}
sessionInfo()
```
